---
title: "Session 0 Practical: Introduction to Stan and cmdstanr"
subtitle: "Bayesian Joint Modelling Training 2026"
author: "Hillary Katsabola"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: flatly
    highlight-style: github
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
---

## Overview

In this practical you will fit seven progressively richer Bayesian models in Stan,
all framed around HIV clinical questions. By the end you will be able to:

1. Fit a **Binomial model** (viral suppression trial) — single-parameter posterior
2. Fit a **Normal model (two unknowns)** (CD4 recovery — mean and SD) — joint posterior, PPC
3. Fit a **Simple Linear Regression** (CD4 count vs ART duration) — slope, PPC, LOO
4. Fit a **Multiple Linear Regression** (ART adherence vs sex + CD4) — multi-predictor, PPC
5. Fit a **Binomial Logistic Regression** (suppression counts per patient) — logit link, Binomial likelihood, PPC on proportions
6. Fit a **Logistic Regression** (per-visit binary suppression) — Bernoulli likelihood, counselling effect, grouped PPC
7. Fit a **Poisson Regression** (clinic visit counts) — log link, rootogram PPC, overdispersion check

All Stan model files live in `stan/`. Data for Models 1–4 are in `data/`.
Data for Models 5–7 are read from `data/` CSV files.

---

## Setup

```{r setup}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(tidyverse)

# Colour palette for plots (SU brand approximation)
color_scheme_set("teal")

# Verify Stan installation
check_cmdstan_toolchain(quiet = TRUE)
cat("CmdStan version:", cmdstan_version(), "\n")
```

---

## Part 1: Binomial Model — Viral Suppression Trial

### Clinical question

In a small pilot trial, $N = 30$ HIV patients start a new first-line ART regimen.
After 12 months, $y = 20$ achieve viral suppression (VL $< 1000$ copies/mL).
What is the probability $\theta$ of suppression on this regimen?

**Model:**
$$y \mid \theta \sim \text{Binomial}(N,\, \theta), \qquad \theta \sim \text{Beta}(2, 2)$$

The Beta(2,2) prior is weakly informative — it places most mass away from 0 and 1
while remaining sceptical of extreme suppression rates.

### Stan model

The model lives in `stan/binomial_coin.stan`:

```stan
// stan/binomial_coin.stan
data {
  int<lower=0> N;                // number of patients
  int<lower=0> n_suppressed;     // number with viral suppression at 12 months
}
parameters {
  real<lower=0, upper=1> theta;   // suppression probability
}
model {
  theta        ~ beta(2, 2);
  n_suppressed ~ binomial(N, theta);
}
generated quantities {
  int y_rep = binomial_rng(N, theta);
}
```

### Compile and fit

```{r binomial-fit}
mod_bin <- cmdstan_model("../stan/binomial_coin.stan")

fit_bin <- mod_bin$sample(
  data    = list(N = 30L, n_suppressed = 20L),
  seed    = 42,
  chains  = 4,
  refresh = 0
)
```

### Summarise

```{r binomial-summary}
fit_bin$summary("theta")
```

**Question 1.1:** What is the posterior mean of $\theta$?
Is there strong evidence that suppression exceeds 50%?

**Question 1.2:** Report the 90% credible interval. Does it include 0.5?

### Visualise the posterior

```{r binomial-plot}
draws_bin <- fit_bin$draws()

mcmc_areas(draws_bin, pars = "theta", prob = 0.90) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "red") +
  labs(title    = "Posterior of \u03b8: Viral Suppression Probability",
       subtitle = "Dashed line = 50% suppression rate",
       x = expression(theta))
```

### Posterior predictive check

```{r binomial-ppc}
y_rep_bin <- as.vector(fit_bin$draws("y_rep", format = "matrix"))

tibble(y_rep = y_rep_bin) |>
  ggplot(aes(x = y_rep)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 20, colour = "red", linewidth = 1.2) +
  labs(title    = "Posterior Predictive Check: Binomial",
       subtitle = "Red line = observed y = 20 suppressed",
       x = "y_rep (simulated suppressed patients out of 30)",
       y = "Count")
```

**Question 1.3:** Does $y = 20$ fall in the bulk of the predictive distribution?

---

## Part 2: Normal Model — Two Unknown Parameters (CD4 Recovery)

### Clinical question

After starting ART, CD4 counts recover. We want to estimate the **mean CD4 recovery**
($\mu$) and its **patient-to-patient variability** ($\sigma$) at 6 months, without
any covariate. This is the two-parameter Normal model from Matsuura (2022, Chapter 3,
extended to two unknowns).

**Model:**
$$Y_n \sim \mathcal{N}(\mu,\, \sigma), \qquad n = 1, \ldots, N$$

Both $\mu$ and $\sigma$ are unknown. In Stan, $\sigma$ is constrained to be positive
via `real<lower=0> sigma` in the `parameters` block.

**Stan model:** `stan/model_normal2.stan`

### Simulate data

We generate $N = 20$ synthetic CD4/100 values representing a realistic 6-month
post-ART recovery distribution ($\mu \approx 0.9$, $\sigma \approx 0.3$):

```{r norm2-data}
set.seed(42)
N_n2   <- 20L
cd4_6m <- rnorm(N_n2, mean = 0.9, sd = 0.3)

cat("N patients:", N_n2, "\n")
cat("Mean CD4/100:", round(mean(cd4_6m), 3), "\n")
cat("SD CD4/100:", round(sd(cd4_6m), 3), "\n")

tibble(cd4 = cd4_6m) |>
  ggplot(aes(x = cd4)) +
  geom_histogram(bins = 10, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = mean(cd4_6m), colour = "red", linewidth = 1) +
  labs(title    = "CD4/100 at 6 Months Post-ART",
       subtitle = "N = 20 simulated patients; red line = sample mean",
       x = "CD4 count / 100 (cells/100 \u03bcL)",
       y = "Count")
```

### Compile and fit

```{r norm2-compile}
mod_n2 <- cmdstan_model("../stan/model_normal2.stan")
```

```{r norm2-fit}
stan_n2 <- list(N = N_n2, cd4_6m = cd4_6m)

fit_n2 <- mod_n2$sample(
  data    = stan_n2,
  seed    = 123,
  chains  = 4,
  refresh = 0
)

fit_n2$summary(c("mu", "sigma"))
```

### Diagnostics

```{r norm2-diag}
fit_n2$diagnostic_summary()

mcmc_trace(fit_n2$draws(), pars = c("mu", "sigma")) +
  labs(title = "Trace Plots — Normal Two-Parameter Model")

mcmc_rhat(fit_n2$summary()$rhat) +
  labs(title = expression(hat(R) ~ "— Normal Two-Parameter Model"))
```

**Question 2.1:** Are all $\hat{R} < 1.01$? Are there any divergent transitions?

**Question 2.2:** What is the posterior mean and 90% credible interval for $\mu$?
How does it compare to the sample mean `mean(cd4_6m)`?

### Joint posterior

The MCMC sample lives in 2D $(\mu, \sigma)$ space. Visualise the joint distribution:

```{r norm2-joint}
mcmc_scatter(fit_n2$draws(), pars = c("mu", "sigma"), alpha = 0.3) +
  labs(title    = "Joint posterior: mean and SD of CD4 recovery",
       x = expression(mu ~ "(mean CD4/100)"),
       y = expression(sigma ~ "(SD CD4/100)"))

mcmc_areas(fit_n2$draws(), pars = c("mu", "sigma"), prob = 0.90) +
  labs(title = "Marginal posteriors with 90% credible intervals")
```

**Question 2.3:** Is there a correlation between $\mu$ and $\sigma$ in the joint
posterior? Why or why not?

### Posterior predictive check

```{r norm2-ppc}
y_rep_n2 <- fit_n2$draws("y_rep", format = "matrix")

ppc_dens_overlay(cd4_6m, y_rep_n2[1:50, ]) +
  labs(title    = "PPC: Normal Two-Parameter — CD4/100 at 6 Months",
       subtitle = "Black = observed; coloured = posterior predictive",
       x = "CD4 count / 100 (cells/100 \u03bcL)")
```

**Question 2.4:** Does the posterior predictive distribution cover the observed data
well? What would a bad PPC look like for this model?

### LOO cross-validation

```{r norm2-loo}
log_lik_n2 <- fit_n2$draws("log_lik", format = "matrix")
loo_n2     <- loo(log_lik_n2)
print(loo_n2)
```

---

## Part 3: Simple Linear Regression — CD4 Count vs ART Duration

### Clinical question

Does CD4 count (cells/$\mu$L) increase with longer ART duration?
We have $N = 15$ HIV patients on long-term ART.

- $X$ = years on antiretroviral therapy (`years_art`)
- $Y$ = CD4 count in cells/100 $\mu$L (multiply $\times 10$ for cells/$\mu$L)

**Model:**
$$Y_i = a + b\, X_i + \varepsilon_i, \qquad \varepsilon_i \sim \text{Normal}(0, \sigma^2)$$

**Stan model:** `stan/model4-4.stan`

### Load and explore data

```{r cd4-data}
cd4 <- read.csv("../data/cd4_art_duration.csv")   # columns: years_art, cd4_per100

cat("Patients:", nrow(cd4), "\n")
summary(cd4)
```

```{r cd4-explore}
ggplot(cd4, aes(x = years_art, y = cd4_per100)) +
  geom_point(size = 2.5, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick", linewidth = 0.8) +
  labs(title    = "CD4 Count vs Years on ART",
       subtitle = "N = 15 HIV patients on long-term ART",
       x = "Years on ART",
       y = "CD4 count (cells / 100 \u03bcL)")
```

### Compile and fit

```{r cd4-compile}
mod_cd4 <- cmdstan_model("../stan/model4-4.stan")
```

```{r cd4-fit}
fit_cd4 <- mod_cd4$sample(
  data    = list(N          = nrow(cd4),
                 years_art  = cd4$years_art,
                 cd4_per100 = cd4$cd4_per100),
  seed    = 1234,
  chains  = 4,
  refresh = 0
)

fit_cd4$summary(c("a", "b", "sigma"))
```

### Diagnostics

```{r cd4-diag}
fit_cd4$diagnostic_summary()

draws_cd4 <- fit_cd4$draws()

mcmc_trace(draws_cd4, pars = c("a", "b", "sigma")) +
  labs(title = "Trace Plots — Simple LR: CD4 vs ART Duration")

mcmc_rhat(fit_cd4$summary()$rhat) +
  labs(title = expression(hat(R) ~ "values (target: all < 1.01)"))
```

**Question 3.1:** What is the posterior mean and 95% credible interval for $b$
(CD4 change per additional year on ART)?
Is there evidence of a positive trend?

**Question 3.2:** Are there any convergence issues?
Report $\hat{R}$ and ESS for $a$, $b$, $\sigma$.

### Posterior distributions

```{r cd4-posterior}
draws_cd4_df <- fit_cd4$draws(format = "df")

draws_cd4_df |>
  select(a, b, sigma) |>
  pivot_longer(everything(), names_to = "param", values_to = "value") |>
  ggplot(aes(x = value, fill = param)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~param, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(title = "Posterior Distributions: Simple LR (CD4 vs ART Duration)",
       x = "Value", y = "Density") +
  theme(legend.position = "none")
```

### Posterior predictive check

The `generated quantities` block in `model4-4.stan` produces `y_rep` directly.

```{r cd4-ppc}
y_rep_cd4 <- fit_cd4$draws("y_rep", format = "matrix")

ppc_dens_overlay(cd4$cd4_per100, y_rep_cd4[1:50, ]) +
  labs(title    = "PPC: Simple LR — CD4 Count",
       subtitle = "Black = observed; coloured lines = posterior predictive draws",
       x = "CD4 count (cells / 100 \u03bcL)")
```

**Question 3.3:** Does the observed CD4 distribution look consistent with the
posterior predictive distribution?

### LOO cross-validation

```{r cd4-loo}
log_lik_cd4 <- fit_cd4$draws("log_lik", format = "matrix")
loo_cd4     <- loo(log_lik_cd4)
print(loo_cd4)
plot(loo_cd4, main = "LOO Pareto-k Diagnostics: Simple LR")
```

**Question 3.4:** Are all Pareto-$k$ values below 0.5? What does this mean for
the reliability of the LOO estimate?

---

## Part 4: Multiple Linear Regression — ART Adherence

### Clinical question

What patient characteristics predict ART adherence?
We have $N = 50$ HIV patients.

- $Y$ = ART adherence proportion (0–1, measured over 6 months)
- Predictors: `sex` (0 = female, 1 = male) and `cd4_init` (CD4 at ART start,
  cells/100 $\mu$L)

**Model:**
$$Y_i = b_1 + b_2\,\text{sex}_i + b_3 \times \frac{\text{cd4\_init}_i}{100} + \varepsilon_i,
\qquad \varepsilon_i \sim \text{Normal}(0,\sigma^2)$$

The CD4 predictor is divided by 100 so the coefficient $b_3$ corresponds to a change
of 100 cells/100 $\mu$L — matching Matsuura (2022, Chapter 5).

**Stan model:** `stan/model5-3.stan`

### Load and explore data

```{r adh-data}
adh <- read.csv("../data/hiv_adherence.csv")   # columns: patient_id, sex, cd4_init, adherence_prop

cat("Patients:", nrow(adh), "\n")
summary(adh)
```

```{r adh-explore}
ggplot(adh, aes(x = cd4_init, y = adherence_prop, colour = factor(sex))) +
  geom_point(size = 2, alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_colour_manual(values  = c("0" = "steelblue", "1" = "tomato"),
                      labels  = c("0" = "Female", "1" = "Male")) +
  labs(title    = "ART Adherence vs Baseline CD4 by Sex",
       subtitle = "N = 50 HIV patients",
       x = "Baseline CD4 (cells / 100 \u03bcL)",
       y = "Adherence proportion",
       colour = "Sex")
```

### Compile and fit

```{r adh-compile}
mod_adh <- cmdstan_model("../stan/model5-3.stan")
```

```{r adh-fit}
fit_adh <- mod_adh$sample(
  data = list(
    N              = nrow(adh),
    sex            = adh$sex,
    cd4_init       = adh$cd4_init / 100,   # per 100 cells/100 µL
    adherence_prop = adh$adherence_prop
  ),
  seed    = 1234,
  chains  = 4,
  refresh = 0
)

fit_adh$summary(c("b[1]", "b[2]", "b[3]", "sigma"))
```

### Diagnostics

```{r adh-diag}
fit_adh$diagnostic_summary()

mcmc_trace(fit_adh$draws(), pars = c("b[1]", "b[2]", "b[3]", "sigma")) +
  labs(title = "Trace Plots — Multiple LR: ART Adherence")
```

**Question 4.1:** What is the direction of $b_2$ (sex effect)?
Does the 95% credible interval exclude zero?
How do you interpret this clinically?

**Question 4.2:** Interpret $b_3$.
What does a one-unit change in `cd4_init/100` correspond to in terms of CD4 cells/$\mu$L?
(Hint: one unit = 100 cells/100$\mu$L = 1000 cells/$\mu$L. Per 10 cells/100$\mu$L the effect is $b_3/10$.)

### Posterior distributions

```{r adh-posterior}
draws_adh <- fit_adh$draws(format = "df")

draws_adh |>
  select(`b[1]`, `b[2]`, `b[3]`) |>
  rename(Intercept = `b[1]`, Sex_male = `b[2]`, CD4_per100 = `b[3]`) |>
  pivot_longer(everything(), names_to = "param", values_to = "value") |>
  ggplot(aes(x = value, fill = param)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~param, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(title = "Posterior Distributions: Multiple LR (ART Adherence)",
       x = "Value", y = "Density") +
  theme(legend.position = "none")
```

### Posterior predictive check

The `model5-3.stan` model outputs `y_rep` in `generated quantities`.

```{r adh-ppc}
y_rep_adh <- fit_adh$draws("y_rep", format = "matrix")

ppc_dens_overlay(adh$adherence_prop, y_rep_adh[1:50, ]) +
  labs(title    = "PPC: Multiple LR — ART Adherence",
       subtitle = "Black = observed; coloured = posterior predictive draws",
       x = "Adherence proportion")

ppc_stat(adh$adherence_prop, y_rep_adh, stat = "mean") +
  labs(title = "PPC: Mean adherence proportion")

ppc_stat(adh$adherence_prop, y_rep_adh, stat = "sd") +
  labs(title = "PPC: SD of adherence proportion")
```

**Question 4.3:** Do the observed summary statistics fall within the posterior
predictive distributions?

---

## Part 5: Binomial Logistic Regression — Suppression Counts per Patient

### Clinical question

Does the per-visit viral suppression rate vary by sex and baseline CD4 count?
We have $N = 50$ HIV patients followed for 6 months. For each patient we record
the number of scheduled clinic visits ($M$) and the number of those visits where
viral load was suppressed ($Y$).

**Model:**
$$Y_n \sim \text{Binomial}(M_n,\, q_n), \qquad
\text{logit}(q_n) = b_1 + b_2\,\text{sex}_n + b_3\,\text{cd4\_init}_n$$

$q_n$ = per-visit suppression probability for patient $n$.
Note: no $\sigma$ parameter — the Binomial distribution fully specifies the
variance via $M_n$ and $q_n$.

**Data:** `data/hiv_clinic_counts.csv`
**Stan model:** `stan/model_binomial_logistic.stan`

### Load and explore data

```{r bin-data}
d_bin <- read.csv("../data/hiv_clinic_counts.csv")   # columns: patient_id, sex, cd4_init, n_visits, n_supp

cat("Patients:", nrow(d_bin), "\n")
cat("n_visits range:", range(d_bin$n_visits), "\n")
cat("n_supp range:", range(d_bin$n_supp), "\n")
cat("Suppression proportions (n_supp/n_visits):\n")
summary(d_bin$n_supp / d_bin$n_visits)
```

```{r bin-explore}
d_bin |>
  mutate(prop_supp = n_supp / n_visits,
         sex_label = ifelse(sex == 1, "Male", "Female")) |>
  ggplot(aes(x = cd4_init / 100, y = prop_supp, colour = sex_label)) +
  geom_point(size = 2.5, alpha = 0.75) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  scale_colour_manual(values = c("Female" = "steelblue", "Male" = "tomato")) +
  labs(title    = "Per-Visit Suppression Rate vs Baseline CD4",
       subtitle = "N = 50 HIV patients; 6-month follow-up",
       x = "Baseline CD4 (cells / 100 \u03bcL)",
       y = "Suppression proportion (Y/M)",
       colour = "Sex")
```

### Compile and fit

```{r bin-compile}
mod_bin2 <- cmdstan_model("../stan/model_binomial_logistic.stan")
```

```{r bin-fit}
stan_bin2 <- list(N        = nrow(d_bin),
                  Sex      = d_bin$sex,
                  cd4_init = d_bin$cd4_init / 100,
                  n_visits = d_bin$n_visits,
                  n_supp   = d_bin$n_supp)

fit_bin2 <- mod_bin2$sample(
  data    = stan_bin2,
  seed    = 123,
  chains  = 4,
  refresh = 0
)

fit_bin2$summary(c("b[1]", "b[2]", "b[3]"))
```

### Diagnostics

```{r bin-diag}
fit_bin2$diagnostic_summary()

mcmc_trace(fit_bin2$draws(), pars = c("b[1]", "b[2]", "b[3]")) +
  labs(title = "Trace Plots — Binomial Logistic Regression")

mcmc_rhat(fit_bin2$summary()$rhat) +
  labs(title = expression(hat(R) ~ "— Binomial Logistic Model"))
```

**Question 5.1:** Are all $\hat{R}$ values below 1.01? Are there any divergent
transitions?

**Question 5.2:** What is the posterior mean and 90% credible interval for
$b[3]$ (CD4 effect)? Is there strong evidence that higher baseline CD4 increases
the per-visit suppression probability?

### Posterior distributions

```{r bin-posterior}
fit_bin2$draws(format = "df") |>
  select(`b[1]`, `b[2]`, `b[3]`) |>
  rename(Intercept = `b[1]`, Sex_male = `b[2]`, CD4_init = `b[3]`) |>
  pivot_longer(everything(), names_to = "param", values_to = "value") |>
  ggplot(aes(x = value, fill = param)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~param, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(title = "Posterior Distributions: Binomial Logistic Regression",
       x = "Value (logit scale)", y = "Density") +
  theme(legend.position = "none")
```

### Posterior predictive check

For Binomial outcomes, compare the observed **suppression proportions** ($Y/M$)
against the posterior predictive proportions.

```{r bin-ppc}
yp_bin2    <- fit_bin2$draws("yp", format = "matrix")
prop_obs   <- d_bin$n_supp / d_bin$n_visits
prop_rep   <- t(apply(yp_bin2, 1, function(r) r / d_bin$n_visits))

ppc_dens_overlay(prop_obs, prop_rep[1:50, ]) +
  labs(title    = "PPC: Binomial Logistic -- Suppression Rate per Patient",
       subtitle = "Black = observed; coloured = posterior predictive draws",
       x = "Suppression proportion (n_supp/n_visits)")

ppc_stat(prop_obs, prop_rep, stat = "mean") +
  labs(title = "PPC: Mean suppression rate")
```

**Question 5.3:** Does the observed distribution of suppression proportions
fall within the posterior predictive distribution? Is the mean well-reproduced?

### LOO cross-validation

```{r bin-loo}
log_lik_bin2 <- fit_bin2$draws("log_lik", format = "matrix")
loo_bin2     <- loo(log_lik_bin2)
print(loo_bin2)
plot(loo_bin2, main = "LOO Pareto-k: Binomial Logistic Model")
```

**Question 5.4:** Are all Pareto-$k$ values below 0.5? Which patients (if any)
are influential observations?

---

## Part 6: Logistic Regression — Per-Visit Binary Suppression

### Clinical question

Does receiving adherence counselling at a clinic visit increase the probability
of viral suppression at that visit? We have visit-level data from HIV patients
(multiple rows per patient).

- `suppressed` ($Y$): viral suppression at that visit (1 = VL $< 1000$, 0 = not)
- `sex`: patient sex (0 = female, 1 = male)
- `cd4_init`: CD4 at ART start (cells/100 $\mu$L)
- `counselling`: received adherence counselling at that visit (0/1)

**Model:**
$$\text{suppressed}[v] \sim \text{Bernoulli}(q[v]), \qquad
q[v] = \text{logit}^{-1}(b_1 + b_2\,\text{sex} + b_3\,\text{cd4\_init} + b_4\,\text{counselling}[v])$$

**Data:** `data/hiv_visit_level.csv`
**Stan model:** `stan/model_logistic.stan`

### Load and explore data

```{r log5-data}
d_log <- read.csv("../data/hiv_visit_level.csv")   # columns: patient_id, sex, cd4_init, counselling, suppressed

cat("Visit records:", nrow(d_log), "\n")
cat("Suppression rate overall:", round(mean(d_log$suppressed), 3), "\n")
cat("Suppression by counselling:\n")
print(tapply(d_log$suppressed, d_log$counselling, mean))
```

```{r log5-explore}
d_log |>
  mutate(counselling_label = ifelse(counselling == 1,
                                    "Counselling", "No counselling")) |>
  ggplot(aes(x = cd4_init / 100, y = suppressed,
             colour = counselling_label)) +
  geom_jitter(width = 0, height = 0.04, alpha = 0.6, size = 2) +
  geom_smooth(method = "glm", method.args = list(family = binomial),
              se = FALSE, linewidth = 0.9) +
  scale_colour_manual(values = c("No counselling" = "steelblue",
                                 "Counselling"    = "firebrick")) +
  labs(title    = "Per-Visit Suppression vs CD4 by Counselling",
       subtitle = "Visit-level data; multiple rows per patient",
       x = "Baseline CD4 (cells / 100 \u03bcL)",
       y = "Viral suppression (1 = suppressed)",
       colour = "Counselling")
```

### Compile and fit

```{r log5-compile}
mod_log5 <- cmdstan_model("../stan/model_logistic.stan")
```

```{r log5-fit}
stan_log5 <- list(V           = nrow(d_log),
                  Sex         = d_log$sex,
                  cd4_init    = d_log$cd4_init / 100,
                  counselling = d_log$counselling,
                  suppressed  = d_log$suppressed)

fit_log5 <- mod_log5$sample(
  data    = stan_log5,
  seed    = 123,
  chains  = 4,
  refresh = 0
)

fit_log5$summary(c("b[1]", "b[2]", "b[3]", "b[4]"))
```

### Diagnostics

```{r log5-diag}
fit_log5$diagnostic_summary()

mcmc_trace(fit_log5$draws(), pars = c("b[1]", "b[2]", "b[3]", "b[4]")) +
  labs(title = "Trace Plots — Logistic Regression: Per-Visit Suppression")

mcmc_rhat(fit_log5$summary()$rhat) +
  labs(title = expression(hat(R) ~ "— Logistic Model"))
```

**Question 6.1:** Summarise convergence: are all $\hat{R} < 1.01$ and ESS $> 400$?

**Question 6.2:** What is the posterior mean and 90% credible interval for $b[4]$
(counselling effect)? Exponentiate to get the odds ratio. Is there strong evidence
that counselling increases suppression odds?

### Posterior distributions

```{r log5-posterior}
fit_log5$draws(format = "df") |>
  select(`b[1]`, `b[2]`, `b[3]`, `b[4]`) |>
  rename(Intercept  = `b[1]`,
         Sex_male   = `b[2]`,
         CD4_init   = `b[3]`,
         Counselling = `b[4]`) |>
  pivot_longer(everything(), names_to = "param", values_to = "value") |>
  ggplot(aes(x = value, fill = param)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~param, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(title = "Posterior Distributions: Logistic Regression",
       x = "Value (logit scale)", y = "Density") +
  theme(legend.position = "none")
```

### Posterior predictive check

For binary outcomes, compare the observed and replicated suppression proportions
within each counselling group.

```{r log5-ppc-grouped}
yp_log5 <- fit_log5$draws("yp", format = "matrix")

ppc_stat_grouped(d_log$suppressed, yp_log5,
  group = factor(d_log$counselling,
                 labels = c("No counselling", "Counselling")),
  stat  = "mean") +
  labs(title    = "PPC: Suppression Rate by Counselling Status",
       subtitle = "Posterior predictive distribution vs observed proportion")
```

```{r log5-ppc-overall}
ppc_stat(d_log$suppressed, yp_log5, stat = "mean") +
  labs(title = "PPC: Overall mean per-visit suppression rate")
```

**Question 6.3:** Does the model reproduce the observed suppression rate in
each counselling group? If not, what might be missing from the model?

**Question 6.4:** Repeated visits from the same patient are treated as
independent here. What is the key limitation of this assumption? Which model
extension (covered in Session 1) would address it?

### LOO cross-validation

```{r log5-loo}
log_lik_log5 <- fit_log5$draws("log_lik", format = "matrix")
loo_log5     <- loo(log_lik_log5)
print(loo_log5)
plot(loo_log5, main = "LOO Pareto-k: Logistic Model")
```

---

## Part 7: Poisson Regression — Clinic Visit Counts

### Clinical question

Do patients with higher baseline CD4 attend more or fewer clinic visits?
We use the **same dataset** as Part 5 (`hiv_clinic_counts.csv`), but now the
count of visits ($M$) is the **outcome**, not the denominator.

**Model:**
$$\texttt{n\_visits}_n \sim \text{Poisson}(\lambda_n), \qquad
\log(\lambda_n) = b_1 + b_2\,\text{sex}_n + b_3\,\text{cd4\_init}_n$$

**Data:** `data/hiv_clinic_counts.csv` (same as Part 5; $M$ is now the response)
**Stan model:** `stan/model_poisson.stan`

### Load and explore data

```{r pois6-data}
d_pois <- read.csv("../data/hiv_clinic_counts.csv")   # columns: patient_id, sex, cd4_init, n_visits, n_supp

cat("Patients:", nrow(d_pois), "\n")
cat("Visit count summary:\n")
summary(d_pois$n_visits)
cat("Mean:", round(mean(d_pois$n_visits), 2),
    " Var:", round(var(d_pois$n_visits), 2),
    " (Poisson expects Var ~ Mean)\n")
```

```{r pois6-explore}
d_pois |>
  mutate(sex_label = ifelse(sex == 1, "Male", "Female")) |>
  ggplot(aes(x = cd4_init / 100, y = n_visits, colour = sex_label)) +
  geom_point(size = 2.5, alpha = 0.75) +
  geom_smooth(method = "glm",
              method.args = list(family = poisson),
              se = FALSE, linewidth = 0.8) +
  scale_colour_manual(values = c("Female" = "steelblue", "Male" = "tomato")) +
  labs(title    = "Clinic Visit Count vs Baseline CD4",
       subtitle = "N = 50 HIV patients; 6-month follow-up",
       x = "Baseline CD4 (cells / 100 \u03bcL)",
       y = "Number of clinic visits",
       colour = "Sex")
```

### Compile and fit

```{r pois6-compile}
mod_pois6 <- cmdstan_model("../stan/model_poisson.stan")
```

```{r pois6-fit}
stan_pois6 <- list(N        = nrow(d_pois),
                   Sex      = d_pois$sex,
                   cd4_init = d_pois$cd4_init / 100,
                   n_visits = d_pois$n_visits)

fit_pois6 <- mod_pois6$sample(
  data    = stan_pois6,
  seed    = 123,
  chains  = 4,
  refresh = 0
)

fit_pois6$summary(c("b[1]", "b[2]", "b[3]"))
```

### Diagnostics

```{r pois6-diag}
fit_pois6$diagnostic_summary()

mcmc_trace(fit_pois6$draws(), pars = c("b[1]", "b[2]", "b[3]")) +
  labs(title = "Trace Plots — Poisson Regression: Clinic Visit Counts")

mcmc_rhat(fit_pois6$summary()$rhat) +
  labs(title = expression(hat(R) ~ "— Poisson Model"))
```

**Question 7.1:** What is the posterior mean and 90% credible interval for
$b[3]$ (CD4 effect on visit count)? Exponentiate for the visit-rate ratio. Does
the CI exclude 1?

**Question 7.2:** What is the posterior mean for $b[1]$? Compute $e^{b[1]}$ —
what does this represent clinically?

### Posterior distributions

```{r pois6-posterior}
fit_pois6$draws(format = "df") |>
  select(`b[1]`, `b[2]`, `b[3]`) |>
  rename(Intercept = `b[1]`, Sex_male = `b[2]`, CD4_init = `b[3]`) |>
  pivot_longer(everything(), names_to = "param", values_to = "value") |>
  ggplot(aes(x = value, fill = param)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~param, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(title = "Posterior Distributions: Poisson Regression",
       x = "Value (log scale)", y = "Density") +
  theme(legend.position = "none")
```

### Posterior predictive check (rootogram)

For count data, a **rootogram** is more informative than a density overlay:
it shows the square-root of observed vs predicted frequencies at each integer value.

```{r pois6-ppc}
mp_pois6 <- fit_pois6$draws("mp", format = "matrix")

ppc_rootogram(d_pois$n_visits, mp_pois6, style = "hanging") +
  labs(title    = "PPC Rootogram: Poisson -- Clinic Visit Counts",
       subtitle = "Bars drop from predicted line; deviation from zero = misfit",
       x = "Number of visits",
       y = expression(sqrt(frequency)))

ppc_stat(d_pois$n_visits, mp_pois6, stat = "var") +
  labs(title    = "PPC: Variance of visit counts",
       subtitle = "If obs variance >> predicted -> consider negative-binomial")
```

**Question 7.3:** Does the rootogram show good agreement between observed and
predicted visit-count frequencies?

**Question 7.4:** Compare the observed variance to the predicted variance from
the PPC. If the observed variance is substantially larger than the posterior
predictive variance, what does this suggest about model fit, and what
alternative model would you try?

### LOO cross-validation

```{r pois6-loo}
log_lik_pois6 <- fit_pois6$draws("log_lik", format = "matrix")
loo_pois6     <- loo(log_lik_pois6)
print(loo_pois6)
plot(loo_pois6, main = "LOO Pareto-k: Poisson Model")
```

---

## Summary

| Model | Outcome type | Stan likelihood | Key output | $N$ |
|-------|-------------|-----------------|-----------|-----|
| 1. Binomial | Binary (aggregate) | `binomial()` | $P(\theta > 0.5)$ | 30 |
| 2. Normal (2 params) | Continuous | `normal()` | $\mu$, $\sigma$; joint posterior | 20 |
| 3. Simple LR | Continuous | `normal()` | Slope $b$; PPC | 15 |
| 4. Multiple LR | Continuous | `normal()` | $b_2$ (sex), $b_3$ (CD4) | 50 |
| 5. Binomial Logistic | Count of successes | `binomial()` | Suppression rate; PPC on Y/M | 50 |
| 6. Logistic | Binary (per visit) | `bernoulli()` | Counselling OR; grouped PPC | V visits |
| 7. Poisson | Count | `poisson_log()` | Visit-rate ratio; rootogram | 50 |

**Convergence checklist (all models):**

- [ ] $\hat{R} < 1.01$ for all parameters
- [ ] ESS$_{\text{bulk}} > 400$ and ESS$_{\text{tail}} > 400$
- [ ] Zero divergent transitions
- [ ] E-BFMI $> 0.2$
- [ ] PPC looks reasonable for each outcome type

---

## Session 1 Preview

In the next session we extend to **repeated measurements** from the same patient —
introducing the Linear Mixed Effects model:

$$y_{ij} = (\beta_0 + u_{0i}) + (\beta_1 + u_{1i})\, t_{ij} + \varepsilon_{ij}$$

where $u_{0i}$ and $u_{1i}$ capture patient-specific deviations in baseline level
and rate of change. We will fit three progressively richer Stan models using
PBC bilirubin data, following Sorensen et al.\ (2016).
