---
title: "Session 1 Practical: Bayesian Longitudinal Modelling"
subtitle: "From Fixed Effects to Varying Slopes — Three Models for log(Serum Bilirubin)"
author: "JM Training 2026"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-float: true
    code-fold: false
    code-tools: true
    theme: flatly
    highlight-style: github
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
---

## Overview

In this practical you will build three nested Bayesian models for the PBC
longitudinal data, following the approach in Sorensen, Hohenstein & Kliegl
(2016, *Frontiers in Psychology*), applied here to clinical biomarker data:

| Model | Description | Stan file |
|-------|-------------|-----------|
| **M1** | Fixed effects only — naive linear regression | `lme_m1_fixed.stan` |
| **M2** | Varying intercepts — patient-specific baseline bilirubin | `lme_m2_intercepts.stan` |
| **M3** | Varying intercepts **and** slopes — full LME *(primary model)* | `lme_m3_full.stan` |

**Learning objectives:**

1. Understand why naive regression violates independence for longitudinal data
2. Implement non-centred parameterisations for hierarchical models in Stan
3. Interpret variance components and the intraclass correlation coefficient (ICC)
4. Use the LKJ prior on the correlation matrix of random effects
5. Compare models with LOO cross-validation and posterior predictive checks
6. Appreciate shrinkage as a feature, not a flaw

---

## Setup

```{r setup}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(tidyverse)
library(survival)
library(patchwork)

color_scheme_set("teal")

# Set a consistent ggplot2 theme
theme_set(theme_minimal(base_size = 12))
```

---

## 1. Data Preparation

```{r data-prep}
source("../R/data_prep.R")

# data_prep.R creates:
#   long_data   — tibble with log_bili, time_yrs, subj_idx, and raw variables
#   stan_lme_m1 / stan_lme_m2 / stan_lme_m3  — Stan data lists (all identical)

cat("Longitudinal data summary\n")
cat("  Patients:             ", n_subj, "\n")
cat("  Total observations:   ", stan_lme_m3$N, "\n")
cat("  Mean visits/patient:  ",
    round(stan_lme_m3$N / stan_lme_m3$n_subj, 1), "\n")
cat("  Follow-up range (yr): ",
    round(range(long_data$time_yrs), 2), "\n")
```

### 1.1 Descriptive Statistics

```{r descriptives}
long_data |>
  summarise(
    n_obs       = n(),
    mean_bili   = round(mean(bili), 2),
    median_bili = round(median(bili), 2),
    sd_bili     = round(sd(bili), 2),
    mean_logbili = round(mean(log_bili), 3),
    sd_logbili   = round(sd(log_bili), 3),
    min_t       = round(min(time_yrs), 2),
    max_t       = round(max(time_yrs), 2)
  ) |>
  knitr::kable(caption = "Summary statistics: PBC longitudinal data")
```

```{r visits-per-patient}
visits_per_patient <- long_data |>
  count(subj_idx, name = "n_visits")

ggplot(visits_per_patient, aes(x = n_visits)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  labs(title = "Distribution of Visit Counts per Patient",
       x = "Number of visits", y = "Number of patients")
```

### 1.2 Serum Bilirubin Distribution

```{r bili-dist}
p1 <- ggplot(long_data, aes(x = bili)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  labs(title = "Raw serum bilirubin (mg/dL)",
       x = "Serum bilirubin", y = "Count")

p2 <- ggplot(long_data, aes(x = log_bili)) +
  geom_histogram(bins = 40, fill = "darkgreen", colour = "white") +
  labs(title = "log(serum bilirubin)",
       x = "log(bilirubin)", y = "Count")

p1 + p2 +
  plot_annotation(title = "Bilirubin: raw vs log scale",
                  subtitle = "Log transformation reduces skewness — we model log(bilirubin)")
```

**Question 1.1:** Why is modelling on the log scale preferable here?

### 1.3 Individual Patient Trajectories

```{r traj-plot}
set.seed(42)
sample_ids <- sample(unique(long_data$subj_idx), 40)

long_data |>
  filter(subj_idx %in% sample_ids) |>
  ggplot(aes(x = time_yrs, y = log_bili, group = subj_idx)) +
  geom_line(alpha = 0.3, colour = "steelblue") +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE,
              colour = "darkred", linewidth = 1.5) +
  labs(title = "PBC: Individual log(Bilirubin) Trajectories — 40 Patients",
       subtitle = "Dark red = loess population trend",
       x = "Time since enrolment (years)", y = "log(Serum Bilirubin)")
```

**Question 1.2:** Does the population trend appear linear over time?

**Question 1.3:** Is there substantial between-patient variation at baseline
(at time = 0)? What does this suggest about the need for random intercepts?

**Question 1.4:** Do some patients show clearly different *slopes* compared to others?
What does this suggest about random slopes?

---

## 2. Model 1: Fixed Effects Only (Naive Linear Regression)

### 2.1 Mathematical Model

$$Y_{ij} = \beta_0 + \beta_1 \, t_{ij} + \varepsilon_{ij}, \qquad \varepsilon_{ij} \overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2)$$

**Priors:**

$$\beta_0 \sim \mathcal{N}(0,\, 5^2), \qquad \beta_1 \sim \mathcal{N}(0,\, 2^2), \qquad \sigma \sim t_4^+(0,\, 1)$$

**Assumption (that we know is wrong):** all 1,945 observations are independent —
even repeated measurements from the same patient.

### 2.2 Compile and Fit

```{r m1-fit}
mod_m1 <- cmdstan_model("../stan/lme_m1_fixed.stan")

fit_m1 <- mod_m1$sample(
  data            = stan_lme_m1,   # N, y, t  (no subj_id in M1)
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  refresh         = 500
)
```

### 2.3 Convergence Diagnostics

```{r m1-diag}
fit_m1$diagnostic_summary()
```

```{r m1-summary}
fit_m1$summary(c("beta0", "beta1", "sigma")) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Model 1: Posterior Summary")
```

```{r m1-trace}
mcmc_trace(
  fit_m1$draws(c("beta0", "beta1", "sigma")),
  facet_args = list(ncol = 1)
) +
  labs(title = "Model 1: Trace Plots")
```

### 2.4 Posterior Predictive Check

```{r m1-ppc}
y_rep_m1 <- fit_m1$draws("y_rep", format = "matrix")

ppc_dens_overlay(stan_lme_m1$y, y_rep_m1[1:50, ]) +
  labs(title = "Model 1 PPC: Density Overlay",
       subtitle = "Dark line = observed; light lines = posterior predictive replications")
```

```{r m1-ppc-stats}
ppc_stat(stan_lme_m1$y, y_rep_m1, stat = "mean") +
  labs(title = "M1 PPC: Mean check")

ppc_stat(stan_lme_m1$y, y_rep_m1, stat = "sd") +
  labs(title = "M1 PPC: SD check")
```

**Question 2.1:** Does Model 1 appear to reproduce the marginal distribution
of log(bilirubin)? Does this mean the model is correct?

### 2.5 Residual Autocorrelation — Why M1 Fails

```{r m1-resid-autocorr}
# Compute posterior mean fitted values
mu_m1 <- fit_m1$draws("mu", format = "matrix") |> colMeans()

resid_m1_df <- long_data |>
  mutate(resid_m1 = log_bili - mu_m1) |>
  group_by(subj_idx) |>
  arrange(time_yrs, .by_group = TRUE) |>
  mutate(visit_num = row_number())

# Plot within-patient residuals for 6 patients
resid_m1_df |>
  filter(subj_idx %in% 1:6) |>
  ggplot(aes(x = visit_num, y = resid_m1)) +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "darkred") +
  facet_wrap(~subj_idx, labeller = label_both) +
  labs(title = "Model 1: Within-patient residuals",
       subtitle = "Systematic trends → independence assumption violated",
       x = "Visit number", y = "Residual (observed − fitted)")
```

```{r m1-autocorr-lag1}
# Compute lag-1 autocorrelation of residuals within patients
autocorr_df <- resid_m1_df |>
  group_by(subj_idx) |>
  filter(n() >= 3) |>   # need at least 3 visits for lag-1 autocorr
  summarise(
    ac1 = cor(resid_m1[-n()], resid_m1[-1], use = "complete.obs")
  )

ggplot(autocorr_df, aes(x = ac1)) +
  geom_histogram(bins = 30, fill = "tomato", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Model 1: Lag-1 Autocorrelation of Within-Patient Residuals",
       subtitle = "Positive autocorrelation → independence violated → SEs too small",
       x = "Lag-1 autocorrelation", y = "Number of patients")

cat("Median lag-1 autocorrelation:", round(median(autocorr_df$ac1, na.rm = TRUE), 3), "\n")
```

**Question 2.2:** Is the median within-patient autocorrelation close to zero?
What does a positive autocorrelation imply for the credible intervals of $\beta_1$?

### 2.6 Comparison with Rizopoulos / `lm()`

```{r m1-rizopoulos}
library(lme4)

# Frequentist naive regression (same wrong independence assumption as M1)
lm_m1 <- lm(log_bili ~ time_yrs, data = long_data)

cat("=== lm() estimates (Rizopoulos-style naive regression) ===\n")
coef(summary(lm_m1))

cat("\n=== Side-by-side comparison: lm() vs Stan M1 ===\n")
stan_m1_summ <- fit_m1$summary(c("beta0", "beta1", "sigma")) |>
  select(variable, mean, sd, q5, q95)

comparison_m1 <- tibble(
  Parameter = c("β₀ (intercept)", "β₁ (slope/yr)", "σ (residual SD)"),
  `lm() estimate` = round(c(coef(lm_m1)["(Intercept)"],
                              coef(lm_m1)["time_yrs"],
                              sigma(lm_m1)), 3),
  `Stan M1 median` = round(stan_m1_summ$mean, 3),
  `Stan 90% CI`    = paste0("[", round(stan_m1_summ$q5, 3), ", ",
                                 round(stan_m1_summ$q95, 3), "]")
)
knitr::kable(comparison_m1, caption = "Model 1: lm() vs Stan — both ignore clustering")
```

**Question 2.3:** Do the `lm()` and Stan M1 estimates agree? What does this
confirm about the Stan implementation? Why are both still *wrong* despite their agreement?

---

## 3. Model 2: Varying Intercepts (Random Intercept Model)

### 3.1 Mathematical Model

$$Y_{ij} = (\beta_0 + u_{0i}) + \beta_1 \, t_{ij} + \varepsilon_{ij}$$

**Random effects and priors:**

$$u_{0i} \overset{\text{iid}}{\sim} \mathcal{N}(0,\, \sigma_u^2), \qquad \sigma_u \sim t_4^+(0,\, 1)$$

**Non-centred (Stan) form:** $u_{0i}^* \sim \mathcal{N}(0,1)$, $u_{0i} = \sigma_u \cdot u_{0i}^*$

**Intraclass Correlation Coefficient:**

$$\text{ICC} = \frac{\sigma_u^2}{\sigma_u^2 + \sigma^2}$$

Fraction of total log-bilirubin variance attributable to stable between-patient differences.

### 3.2 Compile and Fit

```{r m2-fit}
mod_m2 <- cmdstan_model("../stan/lme_m2_intercepts.stan")

fit_m2 <- mod_m2$sample(
  data            = stan_lme_m2,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  refresh         = 500
)
```

### 3.3 Convergence Diagnostics

```{r m2-diag}
fit_m2$diagnostic_summary()
```

```{r m2-summary}
fit_m2$summary(c("beta0", "beta1", "sigma_u", "sigma")) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Model 2: Posterior Summary — Population-Level Parameters")
```

```{r m2-trace}
mcmc_trace(
  fit_m2$draws(c("beta0", "beta1", "sigma_u", "sigma")),
  facet_args = list(ncol = 2)
) +
  labs(title = "Model 2: Trace Plots")
```

### 3.4 Posterior Distributions

```{r m2-intervals}
mcmc_areas(
  fit_m2$draws(c("beta0", "beta1", "sigma_u", "sigma")),
  prob = 0.90
) +
  labs(title = "Model 2: 90% Credible Intervals")
```

**Question 3.1:** Is $\hat{\sigma}_u$ larger or smaller than $\hat{\sigma}$?
What does this mean about the sources of variability in log(bilirubin)?

### 3.5 Intraclass Correlation Coefficient (ICC)

```{r m2-icc}
draws_m2 <- fit_m2$draws(format = "df")

icc_draws <- draws_m2 |>
  mutate(ICC = sigma_u^2 / (sigma_u^2 + sigma^2))

# Posterior summary of ICC
cat("Posterior ICC (fraction of variance between patients):\n")
quantile(icc_draws$ICC, c(0.025, 0.25, 0.5, 0.75, 0.975)) |>
  round(3) |>
  print()

# Visualise
ggplot(icc_draws, aes(x = ICC)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = median(icc_draws$ICC),
             colour = "darkred", linewidth = 1) +
  labs(title = "Model 2: Posterior Distribution of ICC",
       subtitle = "ICC = σ²_u / (σ²_u + σ²)",
       x = "Intraclass Correlation Coefficient", y = "Posterior draws")
```

**Question 3.2:** Interpret the posterior median ICC in clinical terms.
Does this justify the use of random effects?

### 3.6 Distribution of Random Intercepts

```{r m2-ranef-dist}
# Posterior mean random intercept for each patient
u_draws_m2 <- fit_m2$draws("u", format = "matrix")
u0_means_m2 <- colMeans(u_draws_m2)

tibble(u0i = u0_means_m2) |>
  ggplot(aes(x = u0i)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "darkred") +
  labs(title = "Model 2: Posterior Mean Random Intercepts — All 312 Patients",
       subtitle = "Does the distribution look approximately normal?",
       x = expression(hat(u)["0i"]), y = "Count")
```

**Question 3.3:** Do the estimated random intercepts appear approximately
normally distributed? Does any patient stand out as an outlier?

### 3.7 Shrinkage Illustration

```{r m2-shrinkage}
# Raw per-patient means (no pooling)
raw_means_df <- long_data |>
  group_by(subj_idx) |>
  summarise(
    raw_mean = mean(log_bili),
    n_visits = n()
  )

beta0_hat_m2 <- mean(fit_m2$draws("beta0", format = "matrix"))

shrink_df <- tibble(
  subj_idx    = 1:n_subj,
  raw_mean    = raw_means_df$raw_mean,
  pooled_mean = beta0_hat_m2 + u0_means_m2,
  n_visits    = raw_means_df$n_visits
)

ggplot(shrink_df, aes(x = raw_mean, y = pooled_mean, colour = n_visits)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_viridis_c(name = "No. visits") +
  labs(
    title = "Shrinkage: Raw Patient Means vs Model 2 Partial-Pooling Estimates",
    subtitle = "Points below the diagonal are shrunk toward the population mean.\nPatients with fewer visits (darker) show more shrinkage.",
    x = "Raw patient mean of log(bilirubin)",
    y = "Model 2 estimate of patient intercept + β₀"
  )
```

**Question 3.4:** Which patients are shrunk the most? Why?
Is this a problem or a desirable property?

### 3.8 Within-Patient Residuals After M2

```{r m2-resid-check}
mu_m2 <- fit_m2$draws("mu", format = "matrix") |> colMeans()

long_data |>
  mutate(resid_m2 = log_bili - mu_m2) |>
  group_by(subj_idx) |>
  arrange(time_yrs, .by_group = TRUE) |>
  mutate(visit_num = row_number()) |>
  filter(subj_idx %in% 1:6) |>
  ggplot(aes(x = visit_num, y = resid_m2)) +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "darkred") +
  facet_wrap(~subj_idx, labeller = label_both) +
  labs(title = "Model 2: Within-patient residuals",
       subtitle = "Are systematic within-patient trends removed?",
       x = "Visit number", y = "Residual")
```

**Question 3.5:** Compare these residual plots to those from Model 1.
Are the within-patient trends removed? Do any patients still show clear slopes?

### 3.9 Posterior Predictive Check

```{r m2-ppc}
y_rep_m2 <- fit_m2$draws("y_rep", format = "matrix")

ppc_dens_overlay(stan_lme_m2$y, y_rep_m2[1:50, ]) +
  labs(title = "Model 2 PPC: Density Overlay")
```

### 3.10 Comparison with Rizopoulos / `lmer()`

```{r m2-rizopoulos}
# lmer() with random intercept only — same structure as M2
lmer_m2 <- lmer(log_bili ~ time_yrs + (1 | id), data = long_data,
                REML = TRUE)

cat("=== lmer() estimates (random intercept only) ===\n")
print(summary(lmer_m2)$coefficients)
cat("\nRandom effects (variance components):\n")
print(as.data.frame(VarCorr(lmer_m2)))

# Build comparison table
stan_m2_summ <- fit_m2$summary(c("beta0", "beta1", "sigma_u", "sigma")) |>
  select(variable, mean, q5, q95)

lmer_fe <- fixef(lmer_m2)
lmer_vc <- as.data.frame(VarCorr(lmer_m2))

comparison_m2 <- tibble(
  Parameter        = c("β₀ (intercept)", "β₁ (slope/yr)",
                        "σ_u (random intercept SD)", "σ (residual SD)"),
  `lmer() REML`    = round(c(lmer_fe["(Intercept)"], lmer_fe["time_yrs"],
                               lmer_vc$sdcor[1], lmer_vc$sdcor[2]), 3),
  `Stan M2 median` = round(stan_m2_summ$mean, 3),
  `Stan 90% CI`    = paste0("[", round(stan_m2_summ$q5, 3), ", ",
                                 round(stan_m2_summ$q95, 3), "]")
)
knitr::kable(comparison_m2,
             caption = "Model 2: lmer() vs Stan — random intercept model")
```

**Question 3.6:** Do the `lmer()` REML estimates and Stan M2 posterior medians agree?
What does Stan add that `lmer()` does not provide?

---

## 4. Model 3: Varying Intercepts and Slopes (Full LME — Primary Model)

### 4.1 Mathematical Model

$$Y_{ij} = (\beta_0 + u_{0i}) + (\beta_1 + u_{1i})\, t_{ij} + \varepsilon_{ij}$$

**Bivariate random effects:**

$$\begin{pmatrix} u_{0i} \\ u_{1i} \end{pmatrix} \sim \mathcal{N}_2\!\left(\mathbf{0},\; \boldsymbol\Sigma_u\right), \qquad \boldsymbol\Sigma_u = \begin{pmatrix} \sigma_{u_0}^2 & \rho_u\,\sigma_{u_0}\sigma_{u_1} \\ \rho_u\,\sigma_{u_0}\sigma_{u_1} & \sigma_{u_1}^2 \end{pmatrix}$$

**Priors (Sorensen parameterisation):**

$$\sigma_{u_0},\,\sigma_{u_1} \sim t_4^+(0,\,1), \qquad \boldsymbol\Omega_u \sim \text{LKJ}(2), \qquad \sigma \sim t_4^+(0,\,1)$$

**Non-centred (Cholesky) form in Stan:**
$\mathbf{z}_{u,i} \sim \mathcal{N}_2(\mathbf{0}, \mathbf{I})$,
$\mathbf{u}_i = \text{diag}(\boldsymbol\sigma_u)\,\mathbf{L}_u\,\mathbf{z}_{u,i}$
where $\mathbf{L}_u$ is the Cholesky factor of $\boldsymbol\Omega_u$.

### 4.2 Compile and Fit

```{r m3-fit}
mod_m3 <- cmdstan_model("../stan/lme_m3_full.stan")

fit_m3 <- mod_m3$sample(
  data            = stan_lme_m3,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  adapt_delta     = 0.95,   # increase to 0.99 if you see divergences
  refresh         = 500
)
```

### 4.3 Convergence Diagnostics

```{r m3-diag}
fit_m3$diagnostic_summary()
```

```{r m3-summary}
fit_m3$summary(
  c("beta0", "beta1", "sigma_u[1]", "sigma_u[2]", "rho_u", "sigma")
) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(
    variable = c("β₀ (intercept)", "β₁ (slope/yr)",
                 "σ_{u0} (SD intercepts)", "σ_{u1} (SD slopes)",
                 "ρ_u (int–slope corr)", "σ (residual SD)"),
    across(where(is.numeric), \(x) round(x, 3))
  ) |>
  knitr::kable(caption = "Model 3: Posterior Summary — Population-Level Parameters")
```

### 4.4 Trace Plots

```{r m3-trace}
mcmc_trace(
  fit_m3$draws(c("beta0", "beta1", "sigma_u[1]", "sigma_u[2]", "rho_u")),
  facet_args = list(ncol = 2)
) +
  labs(title = "Model 3: Trace Plots — Population-Level Parameters")
```

**Question 4.1:** Does `rho_u` mix as well as the other parameters?
Why might the intercept–slope correlation be harder to estimate?

### 4.5 Posterior Distributions

```{r m3-posterior}
mcmc_areas(
  fit_m3$draws(c("beta0", "beta1", "sigma_u[1]", "sigma_u[2]", "rho_u", "sigma")),
  prob = 0.90, prob_outer = 0.99
) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  labs(title = "Model 3: 90% and 99% Credible Intervals")
```

### 4.6 Interpreting $\rho_u$: The Intercept–Slope Correlation

```{r m3-rho}
rho_draws <- fit_m3$draws("rho_u", format = "matrix") |> as.vector()

cat("Posterior summary of ρ_u:\n")
cat("  Median:", round(median(rho_draws), 3), "\n")
cat("  90% CI:", round(quantile(rho_draws, c(0.05, 0.95)), 3), "\n")
cat("  P(ρ_u > 0):", round(mean(rho_draws > 0), 3), "\n")

ggplot(tibble(rho = rho_draws), aes(x = rho)) +
  geom_histogram(bins = 40, fill = "purple4", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "darkred") +
  geom_vline(xintercept = median(rho_draws), colour = "darkred", linewidth = 1.2) +
  labs(title = "Model 3: Posterior of Intercept–Slope Correlation (ρ_u)",
       subtitle = "Does baseline bilirubin predict the rate of deterioration?",
       x = "ρ_u", y = "Posterior draws")
```

**Question 4.2:** What is the posterior median $\hat{\rho}_u$? Is there
strong evidence it is positive? Interpret this finding clinically: do patients
with higher baseline bilirubin tend to deteriorate faster?

### 4.7 Population-Level Trajectory

```{r m3-population-traj}
draws_m3 <- fit_m3$draws(format = "df")
t_seq    <- seq(0, 14, by = 0.25)

traj_df <- draws_m3 |>
  slice_sample(n = 200) |>
  select(beta0, beta1) |>
  mutate(draw = row_number()) |>
  crossing(t = t_seq) |>
  mutate(mu = beta0 + beta1 * t)

ggplot(traj_df, aes(x = t, y = mu, group = draw)) +
  geom_line(alpha = 0.05, colour = "steelblue") +
  stat_summary(fun = median, geom = "line",
               linewidth = 1.5, colour = "darkred") +
  labs(
    title = "Model 3: Population Trajectory — 200 Posterior Draws",
    subtitle = "Each faint line = one posterior draw; dark red = median",
    x = "Time since enrolment (years)", y = "Mean log(Serum Bilirubin)"
  )
```

### 4.8 Patient-Specific Fitted Trajectories (Shrinkage)

```{r m3-patient-traj}
# Extract posterior mean random effects for all patients
u_m3   <- fit_m3$draws("u", format = "matrix")
u0_bar <- colMeans(u_m3[, grepl("u\\[1,", colnames(u_m3))])
u1_bar <- colMeans(u_m3[, grepl("u\\[2,", colnames(u_m3))])

beta0_hat_m3 <- mean(fit_m3$draws("beta0", format = "matrix"))
beta1_hat_m3 <- mean(fit_m3$draws("beta1", format = "matrix"))

# Patient-specific fitted trajectories for a sample of 20 patients
subj_traj <- tibble(
  subj_idx = 1:n_subj,
  int_i    = beta0_hat_m3 + u0_bar,
  slp_i    = beta1_hat_m3 + u1_bar
) |>
  filter(subj_idx <= 20) |>
  crossing(t = seq(0, 14, by = 0.5)) |>
  mutate(pred = int_i + slp_i * t)

obs_sub20 <- long_data |> filter(subj_idx <= 20)

ggplot() +
  geom_point(data = obs_sub20,
             aes(x = time_yrs, y = log_bili, group = subj_idx),
             alpha = 0.3, size = 0.8, colour = "grey40") +
  geom_line(data = subj_traj,
            aes(x = t, y = pred, group = subj_idx),
            colour = "steelblue", alpha = 0.7, linewidth = 0.8) +
  labs(
    title = "Model 3: Patient-Specific Fitted Trajectories — 20 Patients",
    subtitle = "Blue lines = model-predicted trajectories; dots = observed log(bilirubin)",
    x = "Time since enrolment (years)", y = "log(Serum Bilirubin)"
  )
```

**Question 4.3:** Do the patient-specific trajectories appear to fit the
observed data well? Are there any patients whose trajectories look unexpected?

### 4.9 Distribution of Random Effects

```{r m3-ranef-dist}
# Joint distribution of random intercepts and slopes
ranef_df <- tibble(
  subj_idx = 1:n_subj,
  u0       = u0_bar,
  u1       = u1_bar
)

ggplot(ranef_df, aes(x = u0, y = u1)) +
  geom_point(alpha = 0.5, colour = "purple4") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", se = TRUE, colour = "darkred") +
  labs(
    title = "Model 3: Joint Distribution of Random Intercepts and Slopes",
    subtitle = "Slope of red line reflects ρ_u — positive slope = patients starting high deteriorate faster",
    x = "Random intercept (u₀ᵢ)", y = "Random slope (u₁ᵢ)"
  )
```

**Question 4.4:** Does the scatter plot of $(u_{0i}, u_{1i})$ suggest a positive
correlation? Is this consistent with the posterior of $\rho_u$?

### 4.10 Posterior Predictive Checks

```{r m3-ppc}
y_rep_m3 <- fit_m3$draws("y_rep", format = "matrix")

p1 <- ppc_dens_overlay(stan_lme_m3$y, y_rep_m3[1:50, ]) +
  labs(title = "Density overlay")
p2 <- ppc_stat(stan_lme_m3$y, y_rep_m3, stat = "mean") +
  labs(title = "Mean check")
p3 <- ppc_stat(stan_lme_m3$y, y_rep_m3, stat = "sd") +
  labs(title = "SD check")
p4 <- ppc_stat(stan_lme_m3$y, y_rep_m3, stat = "max") +
  labs(title = "Max check")

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "Model 3: Posterior Predictive Checks")
```

### 4.11 Comparison with Rizopoulos / `lmer()`

```{r m3-rizopoulos}
# lmer() with random intercept AND slope (correlated) — same structure as M3
lmer_m3 <- lmer(log_bili ~ time_yrs + (time_yrs | id), data = long_data,
                REML = TRUE)

cat("=== lmer() estimates (random intercept + slope) ===\n")
print(summary(lmer_m3)$coefficients)
cat("\nRandom effects (variance components):\n")
print(as.data.frame(VarCorr(lmer_m3)))

# Extract Stan M3 posterior summaries
stan_m3_summ <- fit_m3$summary(
  c("beta0", "beta1", "sigma_u[1]", "sigma_u[2]", "rho_u", "sigma")
) |>
  select(variable, mean, q5, q95)

lmer_fe3  <- fixef(lmer_m3)
lmer_vc3  <- as.data.frame(VarCorr(lmer_m3))
lmer_corr <- attr(VarCorr(lmer_m3)$id, "correlation")[2, 1]

comparison_m3 <- tibble(
  Parameter        = c("β₀ (intercept)", "β₁ (slope/yr)",
                        "σ_{u0} (intercept SD)", "σ_{u1} (slope SD)",
                        "ρ_u (int–slope corr)", "σ (residual SD)"),
  `lmer() REML`    = round(c(lmer_fe3["(Intercept)"], lmer_fe3["time_yrs"],
                               lmer_vc3$sdcor[1], lmer_vc3$sdcor[2],
                               lmer_corr, lmer_vc3$sdcor[3]), 3),
  `Stan M3 median` = round(stan_m3_summ$mean, 3),
  `Stan 90% CI`    = paste0("[", round(stan_m3_summ$q5, 3), ", ",
                                 round(stan_m3_summ$q95, 3), "]")
)
knitr::kable(comparison_m3,
             caption = "Model 3: lmer() vs Stan — full LME (Rizopoulos validation)")
```

**Question 4.5:** How closely do the Stan M3 medians match the `lmer()` REML estimates?
What does near-perfect agreement confirm? What can you say about $\rho_u$ from Stan
that you cannot easily quantify from `lmer()`?

---

## 5. Model Comparison via LOO Cross-Validation

### 5.1 Compute LOO Estimates

```{r loo-compute}
library(loo)

ll_m1 <- fit_m1$draws("log_lik", format = "matrix")
ll_m2 <- fit_m2$draws("log_lik", format = "matrix")
ll_m3 <- fit_m3$draws("log_lik", format = "matrix")

loo_m1 <- loo(ll_m1)
loo_m2 <- loo(ll_m2)
loo_m3 <- loo(ll_m3)

cat("=== LOO Summary ===\n")
print(loo_m1)
print(loo_m2)
print(loo_m3)
```

### 5.2 Formal Comparison

```{r loo-compare}
loo_compare(loo_m1, loo_m2, loo_m3)
```

### 5.3 Visual Comparison

```{r loo-plot}
elpd_df <- tibble(
  model = c("M1: Fixed effects",
            "M2: Varying intercepts",
            "M3: Varying int + slopes"),
  elpd  = c(loo_m1$estimates["elpd_loo", "Estimate"],
            loo_m2$estimates["elpd_loo", "Estimate"],
            loo_m3$estimates["elpd_loo", "Estimate"]),
  se    = c(loo_m1$estimates["elpd_loo", "SE"],
            loo_m2$estimates["elpd_loo", "SE"],
            loo_m3$estimates["elpd_loo", "SE"])
)

ggplot(elpd_df, aes(x = reorder(model, elpd), y = elpd)) +
  geom_point(size = 4, colour = "steelblue") +
  geom_errorbar(aes(ymin = elpd - 2*se, ymax = elpd + 2*se),
                width = 0.2, colour = "steelblue") +
  coord_flip() +
  labs(
    title = "LOO-CV: Predictive Accuracy of Three Models",
    subtitle = "Higher ELPD = better out-of-sample prediction (error bars = ±2 SE)",
    x = NULL, y = "ELPD"
  )
```

### 5.4 Pareto-k Diagnostics for Model 3

```{r loo-pareto}
plot(loo_m3, label_points = TRUE)
title("Model 3: Pareto-k Diagnostics\nk > 0.7 = influential observation")
```

**Question 5.1:** Are there any observations with Pareto-k > 0.7 in M3?
If so, which patients do they correspond to, and why might they be influential?

**Question 5.2:** By how many ELPD units does M3 beat M2? Is the difference
more than 2 standard errors (i.e., clearly meaningful)?

**Question 5.3:** Why does M2 already represent a large jump over M1,
but M3 provides a more moderate further improvement?

---

## 6. Side-by-Side Model Summary

### 6.1 Parameter Estimates Comparison

```{r all-models-table}
# Fixed effects
bind_rows(
  fit_m1$summary(c("beta0", "beta1", "sigma")) |>
    mutate(model = "M1: Fixed"),
  fit_m2$summary(c("beta0", "beta1", "sigma")) |>
    mutate(model = "M2: Rand. Intercepts"),
  fit_m3$summary(c("beta0", "beta1", "sigma")) |>
    mutate(model = "M3: Rand. Int+Slopes")
) |>
  select(model, variable, mean, sd, q5, q95) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Fixed Effects Across Models")
```

**Question 6.1:** Does $\hat\beta_1$ (the population slope) change across
models? Should we trust the M1 estimate?

### 6.2 PPC Side-by-Side

```{r ppc-side-by-side}
pa <- ppc_dens_overlay(stan_lme_m1$y, y_rep_m1[1:50, ]) + labs(title = "M1")
pb <- ppc_dens_overlay(stan_lme_m2$y, y_rep_m2[1:50, ]) + labs(title = "M2")
pc <- ppc_dens_overlay(stan_lme_m3$y, y_rep_m3[1:50, ]) + labs(title = "M3")

pa + pb + pc +
  plot_annotation(title = "PPC Comparison: M1, M2, M3")
```

### 6.3 Summary Table

```{r summary-table}
tibble(
  Model        = c("M1: Fixed effects",
                   "M2: Varying intercepts",
                   "M3: Varying int + slopes"),
  `Random effects`   = c("None", "u₀ᵢ ~ N(0, σ²_u0)", "(u₀ᵢ, u₁ᵢ) ~ MVN(0, Σ_u)"),
  `Parameters`       = c("3", "~316", "~630"),
  `LOO rank`         = c("3rd (worst)", "2nd", "1st (best)"),
  `Independence`     = c("Violated", "Improved", "Best"),
  `Use for JM`       = c("No", "Possible", "Yes")
) |>
  knitr::kable(caption = "Three Longitudinal Models — Comparison Summary")
```

---

## 7. The Latent Trajectory for Session 3

Model 3 defines the latent log-bilirubin trajectory for patient $i$:

$$m_i(t) = (\beta_0 + u_{0i}) + (\beta_1 + u_{1i})\,t$$

In Session 3, this trajectory is **linked to the hazard of death/transplant**:

$$h_i(t) = h_0(t)\,\exp\!\bigl(\mathbf{w}_i^\top\boldsymbol\gamma + \alpha_0\,u_{0i} + \alpha_1\,u_{1i}\bigr)$$

The association parameters $\alpha_0$ and $\alpha_1$ quantify:

- $\alpha_0$: does a higher baseline bilirubin level increase the hazard?
- $\alpha_1$: does a steeper bilirubin trajectory increase the hazard?

```{r latent-traj-preview}
# Illustrate the latent trajectory for a few patients
subj_show <- c(1, 5, 10, 20, 50, 100)

traj_show <- tibble(
  subj_idx = subj_show,
  int_i    = beta0_hat_m3 + u0_bar[subj_show],
  slp_i    = beta1_hat_m3 + u1_bar[subj_show]
) |>
  crossing(t = seq(0, 14, by = 0.1)) |>
  mutate(
    m_t      = int_i + slp_i * t,
    label    = paste0("Patient ", subj_idx,
                      "\n(u0=", round(u0_bar[subj_idx], 2),
                      ", u1=", round(u1_bar[subj_idx], 2), ")")
  )

ggplot(traj_show, aes(x = t, y = m_t, colour = factor(subj_idx), group = factor(subj_idx))) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Latent Trajectories m_i(t) for Selected Patients",
    subtitle = "These trajectories will link to survival risk in Session 3",
    x = "Time (years)", y = expression(m[i](t) == log~bilirubin~trajectory),
    colour = "Patient"
  )
```

---

## Summary

| Concept | Key takeaway |
|---------|-------------|
| Within-patient correlation | Ignored by M1 → biased SEs |
| ICC | Quantifies how much of variance is between patients |
| Random intercepts | Capture stable patient-level baselines |
| Non-centred parameterisation | Avoids funnel geometry; essential for hierarchical Stan models |
| LKJ prior | Regularises the correlation between intercepts and slopes |
| Shrinkage | Patients with few observations borrow from the group — reduces prediction error |
| LOO-CV | M3 wins; each added level of random effects improves predictive accuracy |
| Latent trajectory $m_i(t)$ | M3 output fed into the joint model in Session 3 |

**Model 3 is the longitudinal submodel for Session 3:**

$$m_i(t) = (\beta_0 + u_{0i}) + (\beta_1 + u_{1i})\,t$$

The posterior draws of $(u_{0i},\, u_{1i})$ carry uncertainty about each patient's
trajectory into the joint model — this is what makes the joint model fully Bayesian.
