---
title: "Session 2 Practical: Bayesian Survival Analysis"
subtitle: "Three PH Models and Competing Risks Applied to PBC Data"
author: "Hillary Katsabola"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: flatly
    highlight-style: github
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
---

## Overview

In this practical you will:

1. Explore the **PBC survival dataset** (`survival::pbc`) and fit the Kaplan-Meier estimator
2. Fit **three Bayesian parametric PH survival models** in Stan:
   - **Model 1 — Exponential:** constant hazard rate; memoryless baseline
   - **Model 2 — Weibull:** monotone hazard; $\alpha > 1$ gives increasing risk over time
   - **Model 3 — Gompertz:** exponentially changing hazard; $\rho > 0$ means accelerating risk
3. Visualise the **posterior hazard function** for each model
4. Compare models using **LOO-CV** and **Cox-Snell residuals**
5. Extend to **competing risks** (transplant vs.\ death) using cause-specific Gompertz hazards
6. Compare all results against frequentist benchmarks from Rizopoulos (2012)

All three single-event models share the **proportional hazards (PH)** structure:

$$h(t \mid \mathbf{x}_i) = h_0(t)\,\exp(\mathbf{x}_i'\boldsymbol{\gamma})$$

where $h_0(t)$ is the baseline hazard and $\exp(\gamma_k)$ is the **hazard ratio** (HR)
for covariate $k$.  The models differ only in their assumed shape for $h_0(t)$.

> **Why only PH models?**
> Session 3 (joint models) requires evaluating $H_i(T_i)$ analytically at every posterior
> draw.  Exponential, Weibull, and Gompertz all have closed-form $H_0(t)$ — no numerical
> quadrature needed inside Stan.  Gompertz is ultimately chosen for the joint model because
> it additionally captures exponential hazard acceleration.

---

## Setup

```{r setup}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(tidyverse)
library(survival)
library(survminer)
library(flexsurv)
library(patchwork)

color_scheme_set("teal")
```

---

## 1. Data Exploration

```{r data-prep}
source("../R/data_prep.R")

cat("N subjects       :", stan_surv$N, "\n")
cat("Events           :", sum(stan_surv$event), "\n")
cat("Censored         :", sum(stan_surv$event == 0), "\n")
cat("Event rate       :", round(mean(stan_surv$event), 3), "\n")
cat("Follow-up (years):",
    round(min(stan_surv$t_event), 2), "–",
    round(max(stan_surv$t_event), 2), "\n")
```

The **PBC dataset** (Mayo Clinic, 1974–1984) enrolled 418 patients with primary biliary
cirrhosis; 312 were randomised to D-penicillamine or placebo.  The outcome here is the
combined endpoint of death **or** liver transplant (`event = 1`); all remaining patients
are right-censored.

Covariates in the Stan data matrix `X_surv` (columns 1–4):

| Column | Variable | Interpretation |
|--------|----------|----------------|
| 1 | `trt_bin` | Drug (1 = D-penicillamine, 0 = placebo) |
| 2 | `female` | Sex (1 = female) |
| 3 | `age_sc` | Age (scaled by ~10 years) |
| 4 | `stage_num` | Histological stage (1–4) |

### Kaplan-Meier estimate

```{r km-plot}
surv_df <- tibble(time = stan_surv$t_event, event = stan_surv$event)
km_fit  <- survfit(Surv(time, event) ~ 1, data = surv_df)

ggsurvplot(km_fit, data = surv_df,
           conf.int   = TRUE,
           risk.table = TRUE,
           palette    = "darkblue",
           title      = "Kaplan-Meier: PBC Survival (combined endpoint)",
           xlab       = "Time (years)",
           ylab       = "Survival probability")
```

```{r km-summary}
cat("Median survival (KM):", round(summary(km_fit)$table["median"], 2), "years\n")
cat("5-year survival     :", round(summary(km_fit, times = 5)$surv, 3), "\n")
cat("10-year survival    :", round(summary(km_fit, times = 10)$surv, 3), "\n")
```

> **Key observation:** The survival function steepens over time, indicating that the hazard
> **accelerates** with disease duration.  This visual evidence against a constant hazard
> (exponential model) motivates the Weibull and Gompertz alternatives.

**Question 1.1:** At what year does the KM survival probability first drop below 0.5?

**Question 1.2:** Looking at the shape of the KM curve, do you anticipate $\hat\alpha > 1$
(Weibull) and $\hat\rho > 0$ (Gompertz)?  Why?

---

## 2. Model 1: Exponential (Constant Hazard)

The **exponential model** is the simplest parametric survival model.  It assumes the
instantaneous event rate is **constant over time** — the so-called *memoryless* property:

$$h(t \mid \mathbf{x}_i) = \lambda\,\exp(\mathbf{x}_i'\boldsymbol{\gamma}), \qquad
  H(t \mid \mathbf{x}_i) = \lambda\,t\,\exp(\mathbf{x}_i'\boldsymbol{\gamma})$$

Parameters:

- $\lambda > 0$: the baseline event rate (events per year for the reference patient)
- $\boldsymbol{\gamma}$: log-hazard ratios; $\exp(\gamma_k)$ = HR for covariate $k$

**Likelihood contributions:**

| Observation type | Stan `target +=` | Intuition |
|------------------|-----------------|-----------|
| Event ($\delta_i=1$) | $\log\lambda + \mathbf{x}_i'\boldsymbol{\gamma} - H_i(t_i)$ | The density $f(t_i) = h(t_i)\,S(t_i)$ |
| Censored ($\delta_i=0$) | $-H_i(t_i)$ | Survived at least to $t_i$: $S(t_i)$ |

**Priors (weakly informative):**

- $\lambda \sim \text{Gamma}(1,1)$: mean rate = 1 event/year; broad but positive
- $\boldsymbol{\gamma} \sim \mathcal{N}(\mathbf{0}, \mathbf{I})$: regularising; HRs centred at 1 on the HR scale

**Median survival** for the reference patient: $t_{1/2} = \log 2 / \lambda$

### 2.1 Fitting the model

```{r exp-fit}
mod_exp <- cmdstan_model("../stan/surv_exponential.stan")

fit_exp <- mod_exp$sample(
  data            = stan_surv,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  refresh         = 500
)

fit_exp$diagnostic_summary()
```

### 2.2 Posterior Summary

```{r exp-summary}
fit_exp$summary(c("lambda", "gamma", "HR", "t_median_ref")) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Exponential Model: Posterior Summary")
```

**Interpreting the output:**

- `lambda`: posterior mean event rate for the reference patient.  For example,
  $\hat\lambda \approx 0.069$ means roughly **6.9 events per 100 person-years**.
- `HR[1]` (drug): an HR $< 1$ indicates a protective effect of D-penicillamine.
- `HR[3]` (age): an HR $> 1$ means older patients face higher risk per unit of
  standardised age.
- `t_median_ref`: median survival for the reference patient.  For the exponential this
  equals $\log 2 / \lambda$ exactly; you can verify with the formula below.
- `rhat` $< 1.01$ and `ess_bulk` $> 400$ confirm good MCMC convergence.

```{r exp-median-check}
# Verify t_median = log(2) / lambda analytically
lambda_mean <- fit_exp$summary("lambda")$mean
cat("Analytical median (log2 / lambda_mean):", round(log(2) / lambda_mean, 2), "years\n")
cat("Stan t_median_ref posterior mean       :",
    round(fit_exp$summary("t_median_ref")$mean, 2), "years\n")
```

### 2.3 Trace Plots & Convergence

```{r exp-trace}
mcmc_trace(
  fit_exp$draws(c("lambda", "gamma[1]", "gamma[2]")),
  facet_args = list(ncol = 1)
) +
  labs(title    = "Exponential Model: Trace Plots",
       subtitle = "Good mixing: chains should overlap ('hairy caterpillar') with no drift")
```

Healthy MCMC chains show:

1. All four chains exploring the **same parameter region** (complete overlap)
2. **Fast within-chain mixing** (no long runs in one direction — low autocorrelation)
3. **Stationarity**: no upward or downward trend across iterations

### 2.4 Posterior Hazard Function

Since $h(t \mid \mathbf{x}_i) = \lambda\exp(\mathbf{x}_i'\boldsymbol{\gamma})$, the
baseline hazard ($\mathbf{x}_i = \mathbf{0}$) is simply $h_0(t) = \lambda$ — a
horizontal line.  The posterior uncertainty about $\lambda$ translates directly to
uncertainty about the **level** of that flat hazard:

```{r exp-hazard}
lambda_samp <- fit_exp$draws("lambda", format = "matrix") |> as.vector()
t_haz <- seq(0.01, 15, length.out = 150)
set.seed(42)
idx_h <- sample(length(lambda_samp), 400)

haz_exp_df <- map_dfr(idx_h, \(s) tibble(t = t_haz, h = lambda_samp[s])) |>
  group_by(t) |>
  summarise(h_med = median(h),
            h_lo  = quantile(h, 0.05, na.rm = TRUE),
            h_hi  = quantile(h, 0.95, na.rm = TRUE))

ggplot(haz_exp_df, aes(x = t, y = h_med)) +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_ribbon(aes(ymin = h_lo, ymax = h_hi), fill = "steelblue", alpha = 0.25) +
  labs(title    = "Exponential: Posterior Baseline Hazard h₀(t)",
       subtitle = "Constant hazard — the memoryless assumption",
       x = "Time (years)", y = "h(t)") +
  coord_cartesian(ylim = c(0, 0.18))
```

The **flat hazard** is the key assumption: the instantaneous risk of the event is
identical whether you are at year 1 or year 10.  For PBC — a **progressive** disease —
this is biologically questionable.

### 2.5 Posterior Survival Curve

```{r exp-survival}
# Posterior survival S(t) = exp(-lambda * t) for reference patient
surv_exp_df <- map_dfr(idx_h, \(s) {
  tibble(t = t_haz, S = exp(-lambda_samp[s] * t_haz))
}) |>
  group_by(t) |>
  summarise(S_med = median(S),
            S_lo  = quantile(S, 0.05, na.rm = TRUE),
            S_hi  = quantile(S, 0.95, na.rm = TRUE))

# KM reference (used again in Section 5)
km_df <- tibble(t = km_fit$time, S = km_fit$surv,
                S_lo = km_fit$lower, S_hi = km_fit$upper)

ggplot() +
  geom_step(data = km_df, aes(x = t, y = S), linewidth = 1.1, colour = "black") +
  geom_ribbon(data = km_df, aes(x = t, ymin = S_lo, ymax = S_hi),
              alpha = 0.15, fill = "black") +
  geom_line(data = surv_exp_df, aes(x = t, y = S_med),
            colour = "steelblue", linewidth = 1.1) +
  geom_ribbon(data = surv_exp_df, aes(x = t, ymin = S_lo, ymax = S_hi),
              fill = "steelblue", alpha = 0.25) +
  labs(title    = "Exponential: Posterior Survival Curve vs Kaplan-Meier",
       subtitle = "Black = KM (95% CI shaded); Blue = Exponential posterior (reference patient)",
       x = "Time (years)", y = "S(t)") +
  ylim(0, 1) + theme_minimal()
```

### 2.6 Comparison with Rizopoulos / `coxph()`

```{r exp-rizopoulos}
cox_fit <- coxph(
  Surv(time_yrs, event) ~ trt_bin + female + age_sc + stage_num,
  data = surv_data
)

cat("=== Cox PH model (Rizopoulos reference — 4 covariates) ===\n")
print(summary(cox_fit)$coefficients[, c("coef", "exp(coef)", "se(coef)", "Pr(>|z|)")])

stan_exp_hr <- fit_exp$summary("HR") |> select(variable, mean, q5, q95)
cox_hr      <- exp(coef(cox_fit))

comparison_exp <- tibble(
  Covariate     = c("Drug (trt_bin)", "Female", "Age (scaled)", "Stage"),
  `Cox HR`      = round(cox_hr, 3),
  `Stan Exp HR` = round(stan_exp_hr$mean, 3),
  `Stan 90% CI` = paste0("[", round(stan_exp_hr$q5, 3), ", ",
                              round(stan_exp_hr$q95, 3), "]")
)
knitr::kable(comparison_exp,
             caption = "Exponential model vs Cox PH (Rizopoulos comparison)")
```

The Cox PH model is **semi-parametric**: it estimates regression coefficients without
specifying $h_0(t)$.  The Bayesian exponential constrains $h_0(t) = \lambda$ (constant).
Close agreement between the two suggests that the constant-hazard assumption does not
severely bias the HR estimates — but we can do better with a more flexible $h_0(t)$.

**Question 2.1:** Use the formula $t_{1/2} = \log 2 / \lambda$ to verify the `t_median_ref`
posterior mean.  What does this number tell you about a typical PBC patient?

**Question 2.2:** Is the drug HR credibly different from 1.0?  How would you communicate
this uncertainty to a clinician?

**Question 2.3:** What biological property of PBC makes the memoryless (constant hazard)
assumption unrealistic?

---

## 3. Model 2: Weibull (Monotone Hazard)

The **Weibull model** extends the exponential by allowing the hazard to change
*monotonically* over time via a shape parameter $\alpha$:

$$h_0(t) = \alpha\, e^{-\alpha\mu}\, t^{\alpha-1}, \qquad
  H_0(t) = \left(\frac{t}{e^\mu}\right)^\alpha$$

$$h(t \mid \mathbf{x}_i) = h_0(t)\,\exp(\mathbf{x}_i'\boldsymbol{\gamma}), \qquad
  H(t \mid \mathbf{x}_i) = \left(\frac{t}{e^{\mu + \mathbf{x}_i'\boldsymbol{\gamma}/\alpha}}\right)^\alpha$$

**Shape parameter $\alpha$:**

| $\alpha$ | Hazard shape | Typical context |
|----------|-------------|-----------------|
| $\alpha < 1$ | Decreasing | Post-surgical recovery, early mortality |
| $\alpha = 1$ | Constant | Reduces exactly to Exponential |
| $\alpha > 1$ | Increasing | Progressive disease, ageing, cancer |

**Bayesian advantage:** Instead of a Wald test ($H_0: \alpha = 1$), we compute the exact
posterior probability $P(\alpha > 1 \mid \mathbf{y})$ — a direct probabilistic statement
requiring no asymptotic approximation.

**Scale parameter:** $e^\mu$ is the characteristic time of the distribution.
The median survival for the reference patient is $t_{1/2} = e^\mu\cdot(\log 2)^{1/\alpha}$.

**Priors:**

- $\alpha \sim \text{Gamma}(2,1)$: mean 2, standard deviation $\sqrt{2}$; allows both sides of 1
- $\mu \sim \mathcal{N}(2, 4)$: log-scale centred at ~7 years
- $\boldsymbol{\gamma} \sim \mathcal{N}(\mathbf{0}, \mathbf{I})$

### 3.1 Fitting the model

```{r weibull-fit}
mod_wb <- cmdstan_model("../stan/surv_weibull.stan")

fit_wb <- mod_wb$sample(
  data            = stan_surv,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  refresh         = 500
)

fit_wb$diagnostic_summary()
```

### 3.2 Posterior Summary

```{r weibull-summary}
fit_wb$summary(c("alpha", "mu", "HR")) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Weibull Model: Posterior Summary")
```

**Interpreting the output:**

- `alpha`: the shape parameter.  The 90% credible interval `[q5, q95]` is a
  **direct probability statement**: "There is a 90% posterior probability that $\alpha$
  lies in this range given the data."  This is the Bayesian analogue of a confidence
  interval, but with a natural probabilistic interpretation.
- `mu`: the log-scale.  The scale $e^\mu$ controls the time horizon of the distribution.
- `HR[k]`: hazard ratios have the same interpretation as in the exponential model.

```{r weibull-scale}
# Scale parameter and implied median survival
mu_mean    <- fit_wb$summary("mu")$mean
alpha_mean <- fit_wb$summary("alpha")$mean
cat("Scale e^mu (posterior mean) :", round(exp(mu_mean), 2), "years\n")
cat("Implied median survival     :", round(exp(mu_mean) * log(2)^(1/alpha_mean), 2), "years\n")
```

### 3.3 Trace Plots

```{r weibull-trace}
mcmc_trace(
  fit_wb$draws(c("alpha", "mu", "gamma[1]", "gamma[2]")),
  facet_args = list(ncol = 2)
) +
  labs(title    = "Weibull Model: Trace Plots",
       subtitle = "Both shape (alpha) and scale (mu) should converge without correlation")
```

### 3.4 Bayesian Test of Increasing Hazard

```{r weibull-alpha}
alpha_draws <- fit_wb$draws("alpha", format = "matrix") |> as.vector()
mu_draws_wb <- fit_wb$draws("mu",    format = "matrix") |> as.vector()

cat("=== Posterior of alpha (shape) ===\n")
cat("  Mean  :", round(mean(alpha_draws), 3), "\n")
cat("  Median:", round(median(alpha_draws), 3), "\n")
cat("  90% CI:", round(quantile(alpha_draws, c(0.05, 0.95)), 3), "\n")
cat("  P(alpha > 1):", round(mean(alpha_draws > 1), 4), "\n\n")

# Posterior median survival (reference patient)
t_med_wb <- exp(mu_draws_wb) * log(2)^(1 / alpha_draws)
cat("=== Implied median survival (reference patient) ===\n")
cat("  Median:", round(median(t_med_wb), 2), "yr\n")
cat("  90% CI:", round(quantile(t_med_wb, c(0.05, 0.95)), 2), "\n")

ggplot(tibble(alpha = alpha_draws), aes(x = alpha)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "darkred", linewidth = 1.2) +
  annotate("text", x = 1.04, y = Inf, vjust = 2, hjust = 0, size = 4,
           colour = "darkred",
           label = paste0("P(α > 1) = ", round(mean(alpha_draws > 1), 4))) +
  labs(title    = "Weibull: Posterior Distribution of Shape Parameter α",
       subtitle = "Red dashed line = α = 1 (exponential boundary); mass above 1 = evidence for increasing hazard",
       x = expression(alpha ~ "(Weibull shape)"), y = "Count")
```

$P(\alpha > 1 \mid \mathbf{y})$ is the **Bayesian analogue** of a one-sided hypothesis
test for $H_0: \alpha \le 1$.  Unlike a $p$-value, it is directly interpretable:
"Given these data, there is a [probability] chance that the hazard increases monotonically
with time."

### 3.5 Posterior Hazard Function

```{r weibull-hazard}
# Weibull baseline hazard for reference patient (x = 0):
# h_0(t) = alpha * exp(-alpha * mu) * t^(alpha - 1)
t_haz_w <- seq(0.1, 15, length.out = 150)
set.seed(42); idx_w <- sample(length(alpha_draws), 400)

haz_wb_df <- map_dfr(idx_w, \(s) {
  a <- alpha_draws[s]; m <- mu_draws_wb[s]
  tibble(t = t_haz_w,
         h = a * exp(-a * m) * t_haz_w^(a - 1))
}) |>
  group_by(t) |>
  summarise(h_med = median(h),
            h_lo  = quantile(h, 0.05, na.rm = TRUE),
            h_hi  = quantile(h, 0.95, na.rm = TRUE))

ggplot(haz_wb_df, aes(x = t, y = h_med)) +
  geom_line(colour = "darkorange", linewidth = 1.2) +
  geom_ribbon(aes(ymin = h_lo, ymax = h_hi), fill = "darkorange", alpha = 0.25) +
  labs(title    = "Weibull: Posterior Baseline Hazard h₀(t)",
       subtitle = "Monotonically increasing (α > 1): risk grows as a power of time",
       x = "Time (years)", y = "h(t) — instantaneous hazard rate")
```

The **increasing power-law hazard** captures the progressive nature of PBC: risk
accumulates steadily as disease advances.  Compare this to the flat exponential hazard —
the Weibull clearly better reflects the biology.

### 3.6 Comparison with Rizopoulos / `flexsurvreg()`

```{r weibull-rizopoulos}
# flexsurv Weibull (AFT) — same 4 covariates; used to verify alpha
fsurv_wb <- flexsurvreg(
  Surv(time_yrs, event) ~ trt_bin + female + age_sc + stage_num,
  dist = "weibull", data = surv_data
)

cat("=== flexsurv Weibull MLE ===\n")
cat("  Shape (alpha) MLE :", round(exp(coef(fsurv_wb)["shape"]), 3), "\n")
cat("  Stan alpha mean   :", round(fit_wb$summary("alpha")$mean, 3), "\n\n")

# HR comparison: Cox (Rizopoulos) vs Stan Weibull
stan_wb_hr <- fit_wb$summary("HR") |> select(variable, mean, q5, q95)

comparison_wb <- tibble(
  Covariate         = c("Drug (trt_bin)", "Female", "Age (scaled)", "Stage"),
  `Cox HR`          = round(exp(coef(cox_fit)), 3),
  `Stan Weibull HR` = round(stan_wb_hr$mean, 3),
  `Stan 90% CI`     = paste0("[", round(stan_wb_hr$q5, 3), ", ",
                                  round(stan_wb_hr$q95, 3), "]")
)
knitr::kable(comparison_wb,
             caption = "Weibull HRs: Cox PH (Rizopoulos reference) vs Stan Bayesian Weibull")
```

The `flexsurv` MLE of $\alpha$ provides a **frequentist reference point**.
Close agreement between the MLE and the Stan posterior mean confirms that
the prior for $\alpha$ is not dominating the likelihood.

**Question 3.1:** What is $P(\alpha > 1 \mid \mathbf{y})$?  State this as an informal
Bayesian hypothesis test and compare with a frequentist interpretation.

**Question 3.2:** Compute the implied median survival from the posterior of $\alpha$ and
$\mu$.  Is it consistent with the Exponential model's `t_median_ref`?

**Question 3.3:** Looking at the hazard plot, how does $h(t)$ at year 10 compare to year 1?
Express this as a ratio.

---

## 4. Model 3: Gompertz (Exponential Hazard)

The **Gompertz model** allows the hazard to grow or decay *exponentially* over time:

$$h(t \mid \mathbf{x}_i) = \lambda\, e^{\rho t}\, \exp(\mathbf{x}_i'\boldsymbol{\gamma})$$

$$H(t \mid \mathbf{x}_i) = \frac{\lambda}{\rho}\!\left(e^{\rho t} - 1\right)\exp(\mathbf{x}_i'\boldsymbol{\gamma})
  \quad [\rho \ne 0]$$

Parameters:

- $\lambda > 0$: the hazard at time $t = 0$ (initial hazard rate)
- $\rho$: the **hazard growth rate** (unconstrained; positive = growth, negative = decay)
- $\boldsymbol{\gamma}$: log-HRs (identical interpretation to exponential/Weibull)

**Hazard shape by $\rho$:**

| $\rho$ | Shape | Clinical context |
|---------|-------|-----------------|
| $\rho > 0$ | Exponentially increasing | Progressive disease (PBC, cancer, ageing) |
| $\rho = 0$ | Constant | Reduces exactly to Exponential |
| $\rho < 0$ | Exponentially decreasing | Post-surgical benefit, waning risk |

**Hazard doubling time:** When $\rho > 0$, the hazard doubles every $\log 2 / \rho$ years.

**Numerical stability:** The formula $\frac{\lambda}{\rho}(e^{\rho t} - 1)$ is unstable
when $|\rho| \approx 0$.  The Stan code guards this with
`abs(rho) > 1e-6`, falling back to $\lambda t$ (the exponential limit) via L'Hôpital's rule.

**Priors:**

- $\lambda \sim \text{Gamma}(1,1)$: same as exponential
- $\rho \sim \mathcal{N}(0, 1)$: allows both growth ($\rho > 0$) and decay ($\rho < 0$)
- $\boldsymbol{\gamma} \sim \mathcal{N}(\mathbf{0}, \mathbf{I})$

### 4.1 Fitting the model

```{r gompertz-fit}
mod_gz <- cmdstan_model("../stan/surv_gompertz.stan")

fit_gz <- mod_gz$sample(
  data            = stan_surv,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  refresh         = 500
)

fit_gz$diagnostic_summary()
```

### 4.2 Posterior Summary

```{r gompertz-summary}
fit_gz$summary(c("lambda", "rho", "HR", "increasing_hazard")) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Gompertz Model: Posterior Summary")
```

**Interpreting the output:**

- `lambda`: the initial hazard at $t = 0$.  This will be much smaller than the
  exponential model's $\hat\lambda$ because the Gompertz hazard starts low and
  then accelerates.
- `rho`: the growth rate.  If the entire 90% CI is positive, the data strongly support
  increasing hazard.
- `increasing_hazard`: a **binary generated quantity** in Stan — it equals 1 for each
  posterior draw where $\rho > 0$.  Its posterior mean is exactly
  $P(\rho > 0 \mid \mathbf{y})$, the Bayesian posterior probability of increasing hazard.
- `HR[k]`: identical interpretation to Exponential and Weibull.

### 4.3 Trace Plots

```{r gompertz-trace}
mcmc_trace(
  fit_gz$draws(c("lambda", "rho", "gamma[1]", "gamma[2]")),
  facet_args = list(ncol = 2)
) +
  labs(title    = "Gompertz Model: Trace Plots",
       subtitle = "Note: rho is unconstrained (can be negative) — its posterior should be well-identified")
```

### 4.4 Posterior of $\rho$ — Evidence for Increasing Hazard

```{r gompertz-rho}
rho_draws  <- fit_gz$draws("rho",    format = "matrix") |> as.vector()
lam_draws  <- fit_gz$draws("lambda", format = "matrix") |> as.vector()

cat("=== Posterior of rho (hazard growth rate) ===\n")
cat("  Mean  :", round(mean(rho_draws), 3), "\n")
cat("  Median:", round(median(rho_draws), 3), "\n")
cat("  90% CI:", round(quantile(rho_draws, c(0.05, 0.95)), 3), "\n")
cat("  P(rho > 0):", round(mean(rho_draws > 0), 4), "\n")

rho_med <- median(rho_draws)
if (rho_med > 0) {
  cat("  Hazard doubling time (log2 / rho_median):",
      round(log(2) / rho_med, 1), "years\n")
}

mcmc_areas(fit_gz$draws(), pars = "rho", prob = 0.90) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  labs(title    = "Gompertz: Posterior Distribution of ρ (hazard growth rate)",
       subtitle = "ρ > 0 = hazard increases over time; ρ = 0 (red) = exponential boundary")
```

The **hazard doubling time** $\log 2 / \hat\rho$ translates the abstract growth rate
into a clinically meaningful statement: roughly every ___ years, the instantaneous risk
of death or transplant doubles.

### 4.5 Posterior Hazard Function

```{r gompertz-hazard}
# Gompertz baseline hazard for reference patient (x = 0):
# h_0(t) = lambda * exp(rho * t)
t_haz_g <- seq(0, 15, length.out = 150)
set.seed(42); idx_g <- sample(length(rho_draws), 400)

haz_gz_df <- map_dfr(idx_g, \(s) {
  tibble(t = t_haz_g,
         h = lam_draws[s] * exp(rho_draws[s] * t_haz_g))
}) |>
  group_by(t) |>
  summarise(h_med = median(h),
            h_lo  = quantile(h, 0.05, na.rm = TRUE),
            h_hi  = quantile(h, 0.95, na.rm = TRUE))

ggplot(haz_gz_df, aes(x = t, y = h_med)) +
  geom_line(colour = "firebrick", linewidth = 1.2) +
  geom_ribbon(aes(ymin = h_lo, ymax = h_hi), fill = "firebrick", alpha = 0.25) +
  labs(title    = "Gompertz: Posterior Baseline Hazard h₀(t)",
       subtitle = "Exponential acceleration: h(t) = λ·exp(ρ·t)",
       x = "Time (years)", y = "h(t)")
```

The **exponential growth** of the Gompertz hazard contrasts with the Weibull's power-law
growth.  Both are consistent with increasing hazard, but with different curvatures.

### 4.6 Comparison with Rizopoulos / `coxph()` and `flexsurvreg()`

```{r gompertz-rizopoulos}
# cox_fit was defined in Section 2; reused here
cat("=== Cox PH model (Rizopoulos reference) ===\n")
print(summary(cox_fit)$coefficients[, c("coef", "exp(coef)", "se(coef)")])

# flexsurv Gompertz — PH parameterization; exp(coef) gives HR directly
fsurv_gz <- flexsurvreg(
  Surv(time_yrs, event) ~ trt_bin + female + age_sc + stage_num,
  dist = "gompertz", data = surv_data
)

stan_gz_hr <- fit_gz$summary("HR") |> select(variable, mean, q5, q95)

comparison_gz <- tibble(
  Covariate        = c("Drug (trt_bin)", "Female", "Age (scaled)", "Stage"),
  `Rizopoulos Cox` = round(exp(coef(cox_fit)), 3),
  `flexsurv Gomp.` = round(exp(coef(fsurv_gz)[c("trt_bin", "female", "age_sc", "stage_num")]), 3),
  `Stan Gompertz`  = round(stan_gz_hr$mean, 3),
  `Stan 90% CI`    = paste0("[", round(stan_gz_hr$q5, 3), ", ",
                                 round(stan_gz_hr$q95, 3), "]")
)
knitr::kable(comparison_gz,
             caption = "Gompertz HRs: Cox PH (Rizopoulos) vs flexsurv MLE vs Stan Bayesian")
```

**Question 4.1:** What is $P(\rho > 0 \mid \mathbf{y})$?  State this as a Bayesian
posterior probability and interpret it clinically.

**Question 4.2:** Compute the hazard doubling time from the posterior median of $\rho$.
Does this make clinical sense for PBC progression?

**Question 4.3:** Compare `HR[3]` (age) across Exponential, Weibull, and Gompertz models.
Are they consistent?  What does consistency across three different $h_0(t)$ assumptions
tell you?

---

## 5. Model Comparison

Now we formally compare the three PH models using LOO-CV, Cox-Snell residuals,
and survival curve overlays.

### 5.1 LOO-CV

**LOO (Leave-One-Out cross-validation)** estimates out-of-sample predictive accuracy
without refitting.  Higher ELPD (expected log predictive density) is better.
LOOIC $= -2 \times \text{ELPD}$, analogous to AIC.

A $|\Delta\text{ELPD}| > 2 \times \text{SE}$ is considered a meaningful difference.

```{r loo-comparison}
get_ll <- function(fit, var = "log_lik") fit$draws(var, format = "matrix")

loo_exp <- loo(get_ll(fit_exp))
loo_wb  <- loo(get_ll(fit_wb))
loo_gz  <- loo(get_ll(fit_gz))

loo_compare(loo_exp, loo_wb, loo_gz)
```

```{r loo-table}
loo_list <- list(Exponential = loo_exp, Weibull = loo_wb, Gompertz = loo_gz)

loo_table <- map_dfr(names(loo_list), function(nm) {
  lo <- loo_list[[nm]]
  tibble(
    Model = nm,
    ELPD  = round(lo$estimates["elpd_loo", "Estimate"], 1),
    SE    = round(lo$estimates["elpd_loo", "SE"], 1),
    LOOIC = round(-2 * lo$estimates["elpd_loo", "Estimate"], 1)
  )
}) |>
  arrange(desc(ELPD))

knitr::kable(loo_table, caption = "LOO-CV: Three PH Survival Models (higher ELPD = better fit)")
```

### 5.2 Cox-Snell Residuals

**Cox-Snell residuals** $r_i = H(t_i \mid \mathbf{x}_i)$ should follow $\text{Exp}(1)$
under a correctly specified model.  We assess this by plotting the Kaplan-Meier estimate
of the residuals against the $\exp(-r)$ reference line: **good fit = curves coincide**.

```{r cox-snell}
get_cs <- function(fit) colMeans(fit$draws("cox_snell", format = "matrix"))

cs_list <- list(
  Exponential = get_cs(fit_exp),
  Weibull     = get_cs(fit_wb),
  Gompertz    = get_cs(fit_gz)
)

par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 1))
for (nm in names(cs_list)) {
  r     <- cs_list[[nm]]
  km_cs <- survfit(Surv(r, stan_surv$event) ~ 1)
  t_ref <- seq(0, max(r), length.out = 200)
  plot(km_cs, main = nm, xlab = "r = H(t|x)", ylab = "S(r)",
       col = "steelblue", lwd = 2, cex.main = 0.9)
  lines(t_ref, exp(-t_ref), col = "red", lty = 2, lwd = 2)
  legend("topright", c("KM of residuals", "Exp(1) ref"),
         col = c("steelblue", "red"), lty = c(1, 2), cex = 0.7, bty = "n")
}
par(mfrow = c(1, 1))
```

### 5.3 Survival Curve Overlay — All Three Models vs KM

```{r surv-overlay}
# Compute posterior survival S(t) for reference patient from all three models
t_surv <- seq(0, 15, by = 0.25)
set.seed(42)
n_samp  <- 300

# Exponential: S(t) = exp(-lambda * t)
lam_e   <- fit_exp$draws("lambda", format = "matrix") |> as.vector()
idx_e   <- sample(length(lam_e), n_samp)
S_exp_ov <- map_dfr(seq_len(n_samp), \(s) {
  tibble(t = t_surv, model = "Exponential", S = exp(-lam_e[idx_e[s]] * t_surv))
})

# Weibull: S(t) = exp(-(t/scale)^alpha)
draws_wb <- fit_wb$draws(format = "df")
idx_wb   <- sample(nrow(draws_wb), n_samp)
S_wb_ov  <- map_dfr(seq_len(n_samp), \(s) {
  row <- draws_wb[idx_wb[s], ]
  tibble(t = t_surv, model = "Weibull",
         S = pweibull(t_surv, shape = as.numeric(row[["alpha"]]),
                      scale = exp(as.numeric(row[["mu"]])), lower.tail = FALSE))
})

# Gompertz: S(t) = exp(-H(t)), H(t) = (lambda/rho)*(exp(rho*t) - 1)
lam_g  <- fit_gz$draws("lambda", format = "matrix") |> as.vector()
rho_g  <- fit_gz$draws("rho",    format = "matrix") |> as.vector()
idx_gz2 <- sample(length(lam_g), n_samp)
S_gz_ov <- map_dfr(seq_len(n_samp), \(s) {
  l <- lam_g[idx_gz2[s]]; r <- rho_g[idx_gz2[s]]
  H <- if (abs(r) > 1e-6) (l / r) * (exp(r * t_surv) - 1) else l * t_surv
  tibble(t = t_surv, model = "Gompertz", S = exp(-H))
})

S_all <- bind_rows(S_exp_ov, S_wb_ov, S_gz_ov) |>
  group_by(t, model) |>
  summarise(S_med = median(S, na.rm = TRUE),
            S_lo  = quantile(S, 0.05, na.rm = TRUE),
            S_hi  = quantile(S, 0.95, na.rm = TRUE), .groups = "drop")
```

```{r surv-overlay-plot}
pal <- c(Exponential = "steelblue", Weibull = "darkorange", Gompertz = "firebrick")

ggplot() +
  geom_step(data = km_df, aes(x = t, y = S), linewidth = 1.2, colour = "black") +
  geom_ribbon(data = km_df, aes(x = t, ymin = S_lo, ymax = S_hi),
              alpha = 0.15, fill = "black") +
  geom_line(data = S_all, aes(x = t, y = S_med, colour = model), linewidth = 1) +
  geom_ribbon(data = S_all,
              aes(x = t, ymin = S_lo, ymax = S_hi, fill = model),
              alpha = 0.15, colour = NA) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  labs(title    = "Posterior Survival Curves: Three PH Models vs Kaplan-Meier",
       subtitle = "Black = KM (95% CI shaded); coloured = posterior median ± 90% CI (reference patient)",
       x = "Time (years)", y = "S(t)", colour = "", fill = "") +
  ylim(0, 1) + theme_minimal()
```

### 5.4 Hazard Shape Comparison

Overlaying the three posterior hazard functions in a single plot makes the
differences in $h_0(t)$ assumptions immediately visible:

```{r hazard-comparison}
# Bind the three hazard data frames created in sections 2–4
haz_all <- bind_rows(
  haz_exp_df |> mutate(model = "Exponential"),
  haz_wb_df  |> mutate(model = "Weibull"),
  haz_gz_df  |> mutate(model = "Gompertz")
)

ggplot(haz_all, aes(x = t, y = h_med, colour = model)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = h_lo, ymax = h_hi, fill = model),
              alpha = 0.12, colour = NA) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  labs(title    = "Posterior Hazard Functions: Exponential vs Weibull vs Gompertz",
       subtitle = "Reference patient (all covariates = 0). Same data, three different h₀(t) assumptions.",
       x = "Time (years)", y = "h(t) — instantaneous hazard rate",
       colour = "", fill = "") +
  theme_minimal()
```

### 5.5 Hazard Ratio Summary

```{r hr-comparison}
hr_exp <- colMeans(fit_exp$draws("HR", format = "matrix"))
hr_wb  <- colMeans(fit_wb$draws("HR",  format = "matrix"))
hr_gz  <- colMeans(fit_gz$draws("HR",  format = "matrix"))

tibble(
  Covariate   = c("Drug (trt)", "Female", "Age (scaled)", "Stage"),
  `Cox HR`    = round(exp(coef(cox_fit)), 3),
  Exponential = round(hr_exp, 3),
  Weibull     = round(hr_wb,  3),
  Gompertz    = round(hr_gz,  3)
) |>
  knitr::kable(caption = "HR Summary: Cox PH vs Three Bayesian PH Models")
```

**Question 5.1:** Which model wins by LOO-CV?  Is the difference between Weibull and
Gompertz larger than $2 \times \text{SE}$?

**Question 5.2:** Examining the Cox-Snell residual plots, which model's KM curve deviates
most from the $\text{Exp}(1)$ reference?  Which models are hardest to distinguish?

**Question 5.3:** In the survival curve overlay, does the Exponential model over- or
under-estimate survival at early times (< 5 years) compared to the KM?

**Question 5.4:** Looking at the hazard comparison plot, at year 10 the Weibull and
Gompertz hazards are both substantially higher than Exponential.  How does this relate
to the LOO-CV and residual results?

**Question 5.5:** Are the HR estimates consistent across all three models?  What does
consistency of covariate effects across three different baseline hazard shapes imply
about robustness?

---

## 6. Competing Risks — Cause-Specific Hazards

Until now we combined transplant and death into a single event (`event = 1`).
But they represent **biologically distinct processes** — the competing risks framework
models them separately.

### Background

In PBC, the `status` variable has three levels:

| Code | Meaning | Clinical interpretation |
|------|---------|------------------------|
| 0 | Censored | Still alive without transplant at study end |
| 1 | Liver transplant | Treatment — prevents further disease progression |
| 2 | Death | Primary clinical endpoint |

When modelling death, patients who receive a transplant are **not** simply censored.
Their liver disease is removed; they are no longer at risk of death *from PBC* after
transplant.  Treating transplant as censoring overestimates the mortality probability.

The correct framework uses the **Cumulative Incidence Function (CIF)**:

$$F_k(t) = P(T \le t,\;\text{cause} = k) = \int_0^t h_k(s)\,S(s)\,ds$$

where $S(t) = \exp\!\bigl(-H_1(t) - H_2(t)\bigr)$ is the **overall survival** — survival
free of *any* event.  Crucially:

$$F_1(t) + F_2(t) + S(t) = 1 \quad \text{for all } t$$

This means summing the transplant and death CIFs and the overall survival always gives 1
— it is a proper probability partition.

> **Naive mistake to avoid:** Using $1 - \hat{S}(t)$ (from a Kaplan-Meier treating the
> competing event as censored) gives the **cause-specific net probability** — what would
> happen in a world where the competing event cannot occur.  The CIF gives the **crude
> probability** — what actually happens in the real world where both events compete.
> For clinical decision-making, the CIF is almost always what you want.

### Competing risks data

```{r cr-data}
source("../R/data_prep.R")   # stan_competing already available; re-source to be safe

cat("Cause 0 (censored)   :", sum(stan_competing$cause == 0), "\n")
cat("Cause 1 (transplant) :", sum(stan_competing$cause == 1), "\n")
cat("Cause 2 (death)      :", sum(stan_competing$cause == 2), "\n\n")

n_cr <- stan_competing$N
cat("Event rate — transplant:", round(mean(stan_competing$cause == 1), 3), "\n")
cat("Event rate — death     :", round(mean(stan_competing$cause == 2), 3), "\n")
```

### Empirical CIF (Aalen-Johansen estimator)

The **Aalen-Johansen estimator** is the non-parametric estimate of the CIF — analogous
to Kaplan-Meier but accounting for competing events at each failure time.

```{r cr-km}
cr_df <- tibble(
  time  = stan_competing$t_event,
  cause = factor(stan_competing$cause, levels = 0:2,
                 labels = c("censored", "transplant", "death"))
)

aj_fit <- survfit(Surv(time, cause) ~ 1, data = cr_df)

cif_df <- bind_rows(
  tibble(time = aj_fit$time, CIF = aj_fit$pstate[, 2], cause = "Transplant (cause 1)"),
  tibble(time = aj_fit$time, CIF = aj_fit$pstate[, 3], cause = "Death (cause 2)")
)

ggplot(cif_df, aes(x = time, y = CIF, colour = cause)) +
  geom_step(linewidth = 1.1) +
  scale_colour_manual(values = c("Transplant (cause 1)" = "steelblue",
                                 "Death (cause 2)"      = "firebrick")) +
  labs(title    = "Aalen-Johansen Cumulative Incidence Functions",
       subtitle = "F₁(t) + F₂(t) + S(t) = 1 at every time point",
       x = "Time (years)", y = "CIF", colour = "") +
  ylim(0, 0.6) + theme_minimal()
```

**Question 6.1:** At year 10, approximately what fraction of patients have died?  What
fraction has received a transplant?  What fraction is still event-free?  Do these three
fractions add to 1?

**Question 6.2:** Why does neither CIF reach 1.0?  (Hint: there is always a non-zero
probability of remaining event-free.)

---

### Fitting the Bayesian competing risks model

The model uses **Gompertz cause-specific hazards**:
$$h_k(t \mid \mathbf{x}_i) = \lambda_k\, e^{\rho_k t}\,\exp(\mathbf{x}_i'\boldsymbol{\gamma}_k),
\quad k = 1 \text{ (transplant)},\; 2 \text{ (death)}$$

Each cause has its own $\lambda_k$, $\rho_k$, and $\boldsymbol{\gamma}_k$.  The
log-likelihood contributes $-H_1(T_i) - H_2(T_i)$ (overall survival term) to every
observation, plus the density of the observed cause for events.

```{r cr-fit}
mod_cr <- cmdstan_model("../stan/surv_competing_risks.stan")

fit_cr <- mod_cr$sample(
  data            = stan_competing,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1000,
  adapt_delta     = 0.95,
  refresh         = 500
)

fit_cr$diagnostic_summary()
```

### Posterior summary

```{r cr-summary}
fit_cr$summary(c("lambda", "rho", "increasing_hazard")) |>
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Competing Risks: Posterior Summary (baseline hazard per cause)")
```

**Interpreting the output:**

- `lambda[1]`, `lambda[2]`: initial hazard rates for transplant and death respectively.
  A higher $\hat\lambda_2$ means death starts at a higher rate than transplant at $t = 0$.
- `rho[1]`, `rho[2]`: hazard growth rates per cause.  Both positive means both risks
  accelerate over time, though potentially at different rates.
- `increasing_hazard[1]` $= P(\rho_1 > 0 \mid \mathbf{y})$ and
  `increasing_hazard[2]` $= P(\rho_2 > 0 \mid \mathbf{y})$:
  posterior probabilities that each cause's hazard increases.

### Convergence diagnostics

```{r cr-trace}
mcmc_trace(
  fit_cr$draws(c("lambda[1]", "lambda[2]", "rho[1]", "rho[2]")),
  facet_args = list(ncol = 2)
) +
  labs(title    = "Competing Risks: Trace Plots (baseline hazard parameters)",
       subtitle = "Each cause has its own lambda and rho — both should converge independently")
```

### Posterior probability of increasing hazard per cause

```{r cr-increasing}
ih <- fit_cr$draws("increasing_hazard", format = "matrix")
cat("P(rho_1 > 0 | data) — transplant hazard increasing:", round(mean(ih[, 1]), 3), "\n")
cat("P(rho_2 > 0 | data) — death hazard increasing     :", round(mean(ih[, 2]), 3), "\n")
```

**Question 6.3:** Does the transplant hazard increase or decrease with time?
Clinically, would you expect increasing transplant rates over disease duration, or
do most urgent referrals happen at particular disease stages?

### Cause-specific hazard ratios

```{r cr-hr}
hr_draws     <- fit_cr$draws("HR", format = "matrix")
P_cr_hr      <- stan_competing$P
covar_names_cr <- c("trt", "female", "age_per10yr", "stage")

hr_summary <- tibble(
  Covariate = rep(covar_names_cr, 2),
  Cause     = rep(c("Transplant", "Death"), each = P_cr_hr),
  HR_mean   = round(colMeans(hr_draws), 3),
  HR_lo     = round(apply(hr_draws, 2, quantile, 0.05), 3),
  HR_hi     = round(apply(hr_draws, 2, quantile, 0.95), 3)
)

knitr::kable(hr_summary,
             caption = "Cause-Specific Hazard Ratios (posterior mean, 90% CI)")
```

**Question 6.4:** Is the age HR larger for transplant or death?  Does this make clinical
sense — for example, are older patients less likely to be referred for transplant
(higher age HR for death) or more likely to die before getting one?

### Posterior CIF from Stan draws

The posterior CIF is computed by numerically integrating the Gompertz cause-specific
hazards over posterior draws (trapezoidal rule):

```{r cr-cif-posterior}
lam_cr <- fit_cr$draws("lambda", format = "matrix")
rho_cr <- fit_cr$draws("rho",    format = "matrix")

t_grid_cr <- seq(0, 12, length.out = 100)
n_cr_d    <- nrow(lam_cr)
set.seed(42); idx_cif <- sample(n_cr_d, 500)

cif_post <- map_dfr(1:2, function(k) {
  map_dfr(idx_cif, function(s) {
    lam   <- lam_cr[s, k]; rho_k <- rho_cr[s, k]
    H_k   <- if (abs(rho_k) > 1e-7)
               (lam / rho_k) * (exp(rho_k * t_grid_cr) - 1)
             else lam * t_grid_cr
    k2    <- 3 - k
    lam2  <- lam_cr[s, k2]; rho2 <- rho_cr[s, k2]
    H_k2  <- if (abs(rho2) > 1e-7)
               (lam2 / rho2) * (exp(rho2 * t_grid_cr) - 1)
             else lam2 * t_grid_cr
    S_all <- exp(-H_k - H_k2)
    h_k   <- lam * exp(rho_k * t_grid_cr)
    dt    <- c(0, diff(t_grid_cr))
    F_k   <- cumsum(h_k * S_all * dt)
    tibble(t = t_grid_cr, CIF = F_k, cause = k, draw = s)
  })
})

cif_post_sum <- cif_post |>
  group_by(t, cause) |>
  summarise(CIF_med = median(CIF),
            CIF_lo  = quantile(CIF, 0.05),
            CIF_hi  = quantile(CIF, 0.95), .groups = "drop") |>
  mutate(cause_label = if_else(cause == 1, "Transplant", "Death"))

ggplot(cif_post_sum, aes(x = t, y = CIF_med, colour = cause_label,
                          fill = cause_label)) +
  geom_line(linewidth = 1.1) +
  geom_ribbon(aes(ymin = CIF_lo, ymax = CIF_hi), alpha = 0.25, colour = NA) +
  scale_colour_manual(values = c("Transplant" = "steelblue", "Death" = "firebrick")) +
  scale_fill_manual(values   = c("Transplant" = "steelblue", "Death" = "firebrick")) +
  labs(title    = "Posterior Cumulative Incidence Functions",
       subtitle = "Reference patient (all covariates = 0); 90% credible band",
       x = "Time (years)", y = "CIF", colour = "", fill = "") +
  ylim(0, 0.65) + theme_minimal()
```

**Question 6.5:** Do the Bayesian posterior CIF curves broadly agree with the
Aalen-Johansen empirical estimates?  What might cause small discrepancies (consider
the reference patient vs.\ the average patient in the data)?

### Comparison with Rizopoulos / `coxph()` per cause

```{r cr-rizopoulos}
# Build data frame with all 4 covariates from X matrix
pbc_cr_df <- tibble(
  t_event     = stan_competing$t_event,
  cause       = stan_competing$cause,
  trt         = stan_competing$X[, 1],
  female      = stan_competing$X[, 2],
  age_per10yr = stan_competing$X[, 3],
  stage       = stan_competing$X[, 4]
)

# Cause-specific Cox: treat the other cause as censored
cox_cause1 <- coxph(Surv(t_event, cause == 1) ~ trt + female + age_per10yr + stage,
                    data = pbc_cr_df)
cox_cause2 <- coxph(Surv(t_event, cause == 2) ~ trt + female + age_per10yr + stage,
                    data = pbc_cr_df)

cat("=== Cause 1 (Transplant) — cause-specific Cox ===\n")
print(summary(cox_cause1)$coefficients[, c("coef", "exp(coef)", "se(coef)")])
cat("\n=== Cause 2 (Death) — cause-specific Cox ===\n")
print(summary(cox_cause2)$coefficients[, c("coef", "exp(coef)", "se(coef)")])

stan_cr_hr <- fit_cr$summary("HR") |> select(variable, mean, q5, q95)

comparison_cr <- tibble(
  Covariate          = rep(c("Drug (trt)", "Female", "Age per 10yr", "Stage"), 2),
  Cause              = rep(c("Transplant", "Death"), each = 4),
  `Cox HR`           = round(c(exp(coef(cox_cause1)), exp(coef(cox_cause2))), 3),
  `Stan Gompertz HR` = round(stan_cr_hr$mean, 3),
  `Stan 90% CI`      = paste0("[", round(stan_cr_hr$q5, 3), ", ",
                                   round(stan_cr_hr$q95, 3), "]")
)
knitr::kable(comparison_cr,
             caption = "Competing Risks: cause-specific Cox vs Bayesian Gompertz HRs")
```

**Question 6.6:** Does the drug effect differ substantially between the transplant and
death causes?  If D-penicillamine slows disease progression, what direction of effect
on each cause would you predict, and does the data support this?

---

## Summary

| Model | Hazard $h_0(t)$ | Key parameter | Joint model ready? |
|-------|----------------|---------------|-------------------|
| **Exponential** | Constant $\lambda$ | Rate $\lambda > 0$ | Yes |
| **Weibull** | Power-law $\propto t^{\alpha-1}$ | Shape $\alpha$; test $P(\alpha>1)$ | Yes |
| **Gompertz** | Exponential $\lambda e^{\rho t}$ | Growth rate $\rho$; test $P(\rho>0)$ | **Yes — chosen for Session 3** |
| **Competing Risks** | Per-cause Gompertz | $\lambda_k, \rho_k, \boldsymbol{\gamma}_k$ | — |

**Key takeaways:**

1. **All three PH models give consistent HRs** that agree with the Cox semi-parametric
   model — the choice of baseline hazard shape does not substantially distort the
   covariate effect estimates for PBC.

2. **Both Weibull** ($\hat\alpha \approx 1.45$, $P(\alpha>1) > 0.999$) **and Gompertz**
   ($\hat\rho \approx 0.086$, $P(\rho>0) \approx 0.998$) confirm that PBC hazard
   increases with disease duration, consistent with progressive liver damage.

3. **Gompertz is chosen for Session 3** because its cumulative hazard has a closed-form
   solution ($H(t) = \frac{\lambda}{\rho}(e^{\rho t}-1)$), avoiding numerical quadrature
   inside Stan when evaluating the joint model likelihood.

4. **Competing risks** reveal that transplant and death have different covariate profiles —
   a pooled model conflates the two.  The CIF (not $1-S(t)$) is the correct probability
   of each event in the presence of competition.
